---
title: "A Bayesian implementation of the Correct Guess Rate Curve"
subtitle: "Reproduction of Szigeti et al. (2023) and an alternative uncertainty model"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 6, dpi = 110)
options(width = 100)
set.seed(20260721)

# This vignette consumes the installed cgrc.bayes package (build + install with
# `R CMD INSTALL .` from the repo root, or `devtools::install()`), rather than
# sourcing R/ directly. Every cgr_* and nig_draws() call below is the package's.
library(cgrc.bayes)
library(ggplot2)

GRID    <- seq(0, 1, length.out = 101)
N_DRAWS <- 20000
has_jags <- requireNamespace("rjags", quietly = TRUE) &&
            requireNamespace("coda",  quietly = TRUE)

```

# Research question and executive summary

## In plain terms (read this first)

You run a drug trial. Half the people get the real drug, half get a placebo, and
nobody is told which. At the end you ask everyone to guess — and about 65% guess
right. That is a problem: people who *believe* they got the real drug tend to
feel better simply because they believe it, and since more of the drug group
correctly believe they got the drug, this "belief bonus" piles up unevenly on the
drug side. Some of the drug's apparent benefit is really just that unevenness.

The **Correct Guess Rate Curve (CGRC)** asks a what-if: *what would this trial
have shown if only 50% had guessed right* — the coin-flip you'd see if the
blinding had actually worked? You can't rerun the trial, but you can re-weight the
people you already have so the maths reflects 50%-correct guessing, then re-take
the difference between the two groups. That re-weighted difference is the CGRC
estimate.

This document does two things: (1) it **reproduces** that method on a real
microdosing dataset, and (2) it offers a **Bayesian** version that reports the
answer as a *range with a probability attached* ("the effect is 1.1 points, and
there's a 79% chance it's positive") instead of a single number plus a p-value.
It is **not** claimed to be a better answer — it targets the same quantity — just
a more honest way of showing the uncertainty. Along the way it checks the
original paper's arithmetic, finds and fixes a few small errors on both sides,
and flags what still needs the original author. Every number below is computed by
the code shown, not asserted.

## The technical version

Blinding is supposed to distribute expectancy effects equally between arms. If
patients can work out their allocation and expect the active arm to help, that
distribution becomes unequal and the trial tilts toward the active treatment.
Szigeti et al. call this *activated expectancy bias* and propose the Correct
Guess Rate Curve (CGRC): estimate what the trial would have shown had the
correct guess rate been 50%.

This document has **two separate objectives**, kept apart throughout:

1. **Reproduce** the published CGRC estimand and treatment-effect curves
   (Sections 2–4, 8, 10).
2. **Develop an alternative implementation** that propagates uncertainty with a
   posterior rather than by resampling (Sections 5–7, 9, 11).

The Bayesian version is **not** presented as better inference about this trial.
Section 8 shows the original procedure and this one target the same quantity.
The defensible claims are narrower: it removes a Monte Carlo error term with no
scientific content, and it replaces an average of p-values — which has no
accepted interpretation — with a posterior distribution that has one.

**Headline numbers (PANAS, week 1).** Under the CGRC reweighting assumptions,
the estimated contrast at CGR = 0.50 is 1.08 points (95% CrI −1.53 to 3.68),
against an unadjusted 3.16 points at the observed CGR of 0.647. The posterior
probability that the effect is positive falls from ~0.99 to ~0.79.

**What this does not establish.** Attenuation is not a measurement of "the
amount caused by expectancy", and matching the published curve does not
validate the causal assumptions. See Section 12.

---

# The original CGRC procedure

Described accurately before any alternative is offered.

Szigeti's method:

1. Split scores into four strata by treatment × guess.
2. Fit a **kernel density estimate** to each stratum (scikit-learn
   `KernelDensity`, all parameters at default — Gaussian kernel, bandwidth 1.0).
3. Draw samples from the four KDEs in proportions such that the combined
   pseudo-dataset has the **target CGR**.
4. Fit `outcome ~ treatment` to that pseudo-dataset.
5. **Repeat, and average** the resulting estimates and p-values.

Steps 3–5 are the part this document replaces. Steps 1–2 define the estimand
and are retained exactly.

**A note on the resample count.** The paper's text (Fig 2 caption) says the
KDE sampling is "repeated 100 times". The author's source repository
([`szb37/CorrectGuessRateCurve`](https://github.com/szb37/CorrectGuessRateCurve),
`src/config.py`) tells a different story: its **Figure 4 configuration**
(`cgrC_low`) specifies `n_cgrc_trials = 32` over `np.linspace(0, 1, 13)` — 32
resamples across 13 CGR grid points (the options are 32/64/96, never 100). Two
caveats keep this from being airtight: the repo states it reproduces the
*preprint* (a different title, then under peer review), so the config may be
preprint-era rather than what generated the final published figure; and the
Figure-4 block in `run.py` is disabled behind `if False:`. Still, none of the
available options is 100, and if the real count is 32 then — because Monte Carlo
error scales as $1/\sqrt{n_{\text{rep}}}$ — the error is about
$\sqrt{100/32} \approx 1.8\times$ larger than a 100-resample assumption.
Section 8 includes 32 in the ladder so the reader can see it either way.

Two properties of step 5 matter later. Averaging p-values across resamples has
no accepted inferential interpretation — it is not a p-value for any hypothesis
test. And resampling introduces Monte Carlo error into the reported estimate
that is a property of the computation, not of the trial.

---

# The four-stratum estimand

## Plain language

Every participant is in exactly one of four boxes: what they **got** crossed
with what they **guessed**.

| Stratum | Got | Guessed | Guess was |
|---|---|---|---|
| `ACAC` | active | active | correct |
| `ACPL` | active | placebo | wrong |
| `PLAC` | placebo | active | wrong |
| `PLPL` | placebo | placebo | correct |

The **observed CGR** is the share who guessed right:

$$c_{obs} = \frac{n_{ACAC} + n_{PLPL}}{n}$$

Practically: if $c_{obs} = 0.65$, 65% of participants knew what they were
taking, so their expectancies were not balanced across arms.

To ask "what if only 50% had guessed right", we reweight — put more weight on
the wrong-guess boxes and less on the right-guess boxes. But we must not
accidentally change the **arm balance** while doing it. Two ratios pin that down:

$$r = \frac{n_{PLPL}}{n_{PLPL} + n_{ACAC}}, \qquad
  s = \frac{n_{ACPL}}{n_{ACPL} + n_{PLAC}}$$

Practically: $r$ is "of the people who guessed right, what fraction were on
placebo"; $s$ is "of the people who guessed wrong, what fraction were on
active". Holding these fixed means the reweighting only changes *how much
guessing happened*, never *who got what*.

## The four weights

| Weight | Formula | Meaning |
|---|---|---|
| $w_{ACAC}$ | $c(1-r)$ | active, correct guess |
| $w_{PLPL}$ | $cr$ | placebo, correct guess |
| $w_{ACPL}$ | $(1-c)s$ | active, wrong guess |
| $w_{PLAC}$ | $(1-c)(1-s)$ | placebo, wrong guess |

**Algebraically**, the correct-guess weights sum to $c$:

$$w_{ACAC} + w_{PLPL} = c(1-r) + cr = c(1-r+r) = c$$

and the incorrect-guess weights sum to $1-c$:

$$w_{ACPL} + w_{PLAC} = (1-c)s + (1-c)(1-s) = (1-c)(s + 1 - s) = 1-c$$

so all four sum to 1. Section 4 checks this numerically.

## The contrast

$$\Delta(c) =
\underbrace{\frac{w_{ACAC}\mu_{ACAC} + w_{ACPL}\mu_{ACPL}}{w_{ACAC} + w_{ACPL}}}_{\text{active arm}}
-
\underbrace{\frac{w_{PLPL}\mu_{PLPL} + w_{PLAC}\mu_{PLAC}}{w_{PLPL} + w_{PLAC}}}_{\text{placebo arm}}$$

Each arm's mean is a weighted average of its two strata; the denominators
renormalise within arm. $\Delta(0.5)$ is the perfectly blinded estimate.

**The key identity.** At $c = c_{obs}$ the weights are proportional to the
observed stratum shares, so each arm collapses to its ordinary sample mean:

$$\Delta(c_{obs}) = \bar{y}_{AC} - \bar{y}_{PL}$$

This is the single most useful check available. It means the green reference
line must intersect the curve exactly at the unadjusted difference — and it is
how a misplaced reference line gets caught.

---

# Worked calculation by hand

Every number below is reproducible with a calculator.

```{r load-data}
stopifnot(file.exists("data/pacutes.csv"))
raw <- read.csv("data/pacutes.csv", stringsAsFactors = FALSE)

if (requireNamespace("digest", quietly = TRUE)) {
  sha <- digest::digest("data/pacutes.csv", algo = "sha256", file = TRUE)
  stopifnot(identical(
    sha, "86aa784528ee045c61fadf3eacfd3e1897d16aae9839cee7cb4bfe839a7cc4e3"))
}

load_scale <- function(code, tp = "w1s1") {
  d <- raw[raw$test_name == code & raw$tp == tp, ]
  data.frame(condition = ifelse(d$condition == "MD", "AC", "PL"),
             guess     = ifelse(d$guess     == "MD", "AC", "PL"),
             value     = d$value, stringsAsFactors = FALSE)
}

SCALES <- c("PANAS" = "PANAS", "Mood VAS" = "mood",
            "Energy VAS" = "energy", "Cognitive performance" = "CPS")
sb <- lapply(SCALES, load_scale); names(sb) <- names(SCALES)
panas <- sb[["PANAS"]]
```

```{r worked-strata}
st  <- cgr_strata(panas)
rat <- cgr_ratios(st)
obs <- cgr_observed(st)

data.frame(stratum = STRATA,
           n    = unname(lengths(st)[STRATA]),
           mean = round(unname(vapply(st[STRATA], mean, numeric(1))), 4),
           sd   = round(unname(vapply(st[STRATA], sd,   numeric(1))), 4),
           sum  = unname(vapply(st[STRATA], sum, numeric(1))))

```

Expected: ACAC n=48 mean=19.5625; ACPL n=43 mean=12.3256; PLAC n=39
mean=18.3590; PLPL n=102 mean=10.9314.

$$c_{obs} = \frac{48 + 102}{232} = \frac{150}{232} = 0.646552$$
$$r = \frac{102}{102+48} = 0.680000 \qquad s = \frac{43}{43+39} = 0.524390$$

```{r worked-weights}
do.call(rbind, lapply(c(obs, 0.5), function(cc) {
  w <- cgr_weights(cc, rat$r, rat$s)
  data.frame(cgr = round(cc, 6),
             w_ACAC = round(w[["ACAC"]], 6), w_ACPL = round(w[["ACPL"]], 6),
             w_PLAC = round(w[["PLAC"]], 6), w_PLPL = round(w[["PLPL"]], 6),
             correct_sum = round(w[["ACAC"]] + w[["PLPL"]], 6),
             incorrect_sum = round(w[["ACPL"]] + w[["PLAC"]], 6))
}))
```

At $c = c_{obs}$: active arm $= 6.331897/0.392241 = 16.142857$, placebo arm
$= 7.892241/0.607759 = 12.985816$, so $\Delta = 3.157042$.

At $c = 0.50$: active arm $= 6.361707/0.422195 = 15.068169$, placebo arm
$= 8.082520/0.577805 = 13.988321$, so $\Delta = 1.079847$.

```{r worked-identity}
mu <- lapply(st, mean)
data.frame(
  quantity = c("Delta(c_obs)", "raw AC - PL mean difference",
               "Delta(0.50)", "absolute identity error"),
  value = c(cgr_delta(obs, mu, rat$r, rat$s),
            mean(panas$value[panas$condition == "AC"]) -
              mean(panas$value[panas$condition == "PL"]),
            cgr_delta(0.5, mu, rat$r, rat$s),
            abs(cgr_delta(obs, mu, rat$r, rat$s) -
                  (mean(panas$value[panas$condition == "AC"]) -
                     mean(panas$value[panas$condition == "PL"]))))
)

```

The identity holds exactly.

---

# Bayesian probability model

Each stratum $j \in \{ACAC, ACPL, PLAC, PLPL\}$ gets its own mean and variance,
with a Normal-Inverse-Gamma prior:

$$y_{ij} \mid \mu_j, \sigma_j^2 \sim \mathcal{N}(\mu_j, \sigma_j^2)$$
$$\mu_j \mid \sigma_j^2 \sim \mathcal{N}\!\left(m_0, \frac{\sigma_j^2}{k_0}\right)$$
$$\sigma_j^2 \sim \text{Inverse-Gamma}(a_0, b_0)$$

The middle line is the one that matters and the one most easily got wrong: the
prior on $\mu_j$ **depends on** $\sigma_j^2$. That is what makes the model
conjugate. A model with $\mu_j \sim \mathcal{N}(m_0, \text{const})$
independently of $\sigma_j^2$ is a *different model* — see Section 8.

| Parameter | Mathematical role | Practical meaning | Behaviour as $n$ grows |
|---|---|---|---|
| $m_0$ | prior mean of $\mu_j$ | where you'd guess the stratum mean sits before seeing data | influence $\to 0$ |
| $k_0$ | prior precision scale | **prior sample size in pseudo-observations** | swamped by $n$ |
| $a_0$ | IG shape | prior confidence about the variance | swamped by $n/2$ |
| $b_0$ | IG scale | prior scale of the variance | swamped by $SS/2$ |

Defaults are $m_0 = 0$, $k_0 = 10^{-6}$, $a_0 = b_0 = 10^{-3}$: deliberately
vague, so the posterior mean lands on the sample mean and the reproduction check
in Section 10 is like-for-like. $k_0 = 10^{-6}$ literally means the prior is
worth one-millionth of an observation.

---

# Posterior derivation

Conjugacy gives closed-form updates:

$$k_n = k_0 + n$$
$$m_n = \frac{k_0 m_0 + n\bar{y}}{k_n}$$
$$a_n = a_0 + \frac{n}{2}$$
$$b_n = b_0 + \frac{1}{2}\sum_i (y_i - \bar{y})^2 + \frac{k_0 n (\bar{y} - m_0)^2}{2k_n}$$

Reading them:

- $k_n$: total information, prior pseudo-observations plus real ones.
- $m_n$: a **precision-weighted average** of prior mean and sample mean. With
  $k_0$ tiny, it is the sample mean.
- $a_n$: each observation contributes half a unit of shape.
- $b_n$: prior scale, plus observed scatter, plus a penalty for prior/data
  disagreement. That third term vanishes when $k_0 \to 0$.

```{r nig-numbers}
d <- nig_draws(st$ACAC, n_draws = 10, return_sigma2 = TRUE)
h <- d$hyper
data.frame(quantity = c("n", "ybar", "sum of squares", "k_n", "m_n",
                        "a_n", "b_n", "m_n - ybar", "E[sigma2]=b_n/(a_n-1)",
                        "sample variance"),
           value = c(h[["n"]], h[["ybar"]], h[["ss"]], h[["kn"]], h[["mn"]],
                     h[["an"]], h[["bn"]], h[["mn"]] - h[["ybar"]],
                     h[["bn"]] / (h[["an"]] - 1), var(st$ACAC)))
```

$m_n$ differs from $\bar{y}$ by about $4 \times 10^{-7}$. That is what "vague
prior" means operationally.

## Why draw $\sigma^2$ first

The joint posterior factorises as

$$p(\mu, \sigma^2 \mid y) = p(\sigma^2 \mid y)\, p(\mu \mid \sigma^2, y)$$

The marginal $p(\sigma^2 \mid y)$ is Inverse-Gamma$(a_n, b_n)$ in closed form,
and the conditional $p(\mu \mid \sigma^2, y)$ is
$\mathcal{N}(m_n, \sigma^2/k_n)$. Drawing in that order — $\sigma^2$, then
$\mu$ given it — produces exact draws from the joint.

**Two facts that are easy to conflate:**

- Iterations are **independent of each other**. Draw 500 tells you nothing
  about draw 501.
- Within a **single** iteration, $\mu$ and $\sigma^2$ are **dependent**: a draw
  with large $\sigma^2$ is paired with a more dispersed $\mu$.

That pairing is the mechanism by which variance uncertainty propagates into
uncertainty about the mean. Inspect it directly:

```{r paired-draws, fig.height=3.5}
pd <- nig_draws(st$ACAC, n_draws = 4000, return_sigma2 = TRUE)
plot(sqrt(pd$sigma2), pd$mu, pch = 16, cex = 0.3,
     col = grDevices::adjustcolor("black", 0.25),
     xlab = expression(sigma~"(within-draw)"), ylab = expression(mu),
     main = "Paired posterior draws: larger sigma gives more dispersed mu")
abline(h = mean(st$ACAC), col = "red", lty = 2)

```

## `n_draws` is not a sample size

Renamed from `S` to remove the ambiguity. `n_draws = 20000` reduces Monte Carlo
error **in the posterior summaries** to roughly $\text{sd}/\sqrt{20000}$. It
adds no information about participants. The data contain 232 people whatever
`n_draws` is set to.

---

# R implementation

The implementation is the installed **`cgrc.bayes`** package (loaded in the
setup chunk with `library(cgrc.bayes)`), so this document is a vignette rather
than the only copy of the code. The package is built from these modules:

| File | Contents |
|---|---|
| `R/01_estimand.R` | strata, ratios, weights, $\Delta(c)$, analytic curve, `cgr_reference_line_test()` |
| `R/02_bayes.R` | `nig_draws()`, conjugate posterior curve |
| `R/03_jags.R` | corrected JAGS model, backend check |
| `R/04_kde.R` | KDE procedure port; `cgr_kde_curve()` for the published figures |
| `R/05_sim.R` | AEB generative model, operating characteristics |
| `R/06_plot.R` | Bayesian figures, `szigeti_panel()` published-style panels, summary tables |
| `R/07_rope.R` | `cgr_rope()` region-of-practical-equivalence decomposition |

Core functions, unfolded because they are short and load-bearing:

```{r core-code, echo=TRUE, eval=FALSE}
cgr_weights <- function(c, r, s) {
  c(ACAC = c * (1 - r), ACPL = (1 - c) * s,
    PLAC = (1 - c) * (1 - s), PLPL = c * r)
}

cgr_delta <- function(c, mu, r, s) {
  w <- cgr_weights(c, r, s)
  (w[["ACAC"]] * mu$ACAC + w[["ACPL"]] * mu$ACPL) / (w[["ACAC"]] + w[["ACPL"]]) -
    (w[["PLPL"]] * mu$PLPL + w[["PLAC"]] * mu$PLAC) / (w[["PLPL"]] + w[["PLAC"]])
}
```

---

# Validation against the original method

Before any table, the published **figures** need reproducing in their own visual
language, so a reader can lay this work beside the paper. Szigeti's Figures 3 and
4 use a twin axis: **blue averaged p-value** on the left, **red treatment
estimate** on the right, a magenta 0.05 threshold, a green "Original CGR" line,
and a black "True blind CGR" line at 0.50. The p-values are the KDE-resampling
averages (Section 6), *not* posterior quantities — this section deliberately
reproduces the p-value presentation the rest of the document argues against,
because that is what reproduction means.

## Figure 3 reproduced: the AEB simulation

Four scenarios crossing direct treatment effect (DTE) and activated expectancy
bias (AEB). "Original CGR" is the simulated guess probability $p_{CG} = 0.7$;
"True blind CGR" is 0.50. The signature to check against the paper is panel 3
(`DTE off, AEB on`): a rising red estimate crossing zero near 0.50, so the blue
p-value peaks near 0.50 and drops toward significance at both ends.

```{r repro-fig3, cache=TRUE, fig.height=8, fig.width=10}
CFG <- list(c(0, 0), c(1, 0), c(0, 1), c(1, 1))
labs <- c("DTE off; AEB off", "DTE on; AEB off",
          "DTE off; AEB on", "DTE on; AEB on")
set.seed(303)
par(mfrow = c(2, 2))
for (i in seq_along(CFG)) {
  d <- sim_aeb(230, 0.7, as.logical(CFG[[i]][1]), as.logical(CFG[[i]][2]), "all")
  szigeti_panel(cgr_kde_curve(d), labs[i], orig_cgr = 0.7, legend = (i == 1))
}
par(mfrow = c(1, 1))
```

## Figure 4 reproduced: the microdose trial

The empirical panels. The green line is drawn at **0.72** because that is where
the published figure and the author's code (`trial_cgrs = {'sbmd': 0.72}`) put
it — the reference-line test below shows why that position is worth checking.
Cognitive performance is the negative control: flat in both curves, because it is
measured by objective tasks rather than self-report, so it should carry no
expectancy signal.

```{r repro-fig4, cache=TRUE, fig.height=8, fig.width=10}
set.seed(404)
par(mfrow = c(2, 2))
for (i in seq_along(sb)) {
  szigeti_panel(cgr_kde_curve(sb[[i]]), names(sb)[i],
                orig_cgr = 0.72, legend = (i == 1))
}
par(mfrow = c(1, 1))
```

These should match the published Figures 3 and 4 panel for panel. The tables that
follow quantify what the eye sees.

## Stratum allocation, supplementary example

```{r val-alloc}
ex <- data.frame(
  condition = c(rep("PL", 65), rep("AC", 35), rep("PL", 25), rep("AC", 75)),
  guess     = c(rep("PL", 65), rep("PL", 35), rep("AC", 25), rep("AC", 75)),
  value = 0)
rbind(
  data.frame(source = "supplementary target", ACAC = 54, ACPL = 58,
             PLAC = 42, PLPL = 46),
  data.frame(source = "this implementation",
             as.list(cgr_sizes(cgr_strata(ex), 0.5))),
  data.frame(source = "shipped-code 2dp rounding",
             as.list(cgr_sizes(cgr_strata(ex), 0.5, legacy_round = TRUE))))

```

## The observed-CGR identity across all outcomes

```{r val-identity}
do.call(rbind, lapply(names(sb), function(nm) {
  d <- sb[[nm]]; s <- cgr_strata(d); rr <- cgr_ratios(s)
  g <- cgr_delta(cgr_observed(s), lapply(s, mean), rr$r, rr$s)
  w <- mean(d$value[d$condition == "AC"]) - mean(d$value[d$condition == "PL"])
  data.frame(outcome = nm, n = nrow(d), delta_at_obs = g, raw_diff = w,
             abs_error = abs(g - w), pass = abs(g - w) < 1e-10)
}))
```

## Faithful reproduction: Szigeti rounds the ratios

The identity above uses the **exact** within-class ratios $r$ and $s$. The
author's code does not: `get_strata_ratio` in
[`szb37/CorrectGuessRateCurve`](https://github.com/szb37/CorrectGuessRateCurve)
(`src/cgrc/core.py`) computes each stratum proportion as `round(x, 2)` before
forming $r$ and $s$. So the **faithful reproduction path is
`legacy_round = TRUE`**, and the exact-ratio default used elsewhere in this
document — while mathematically the cleaner estimand, and the one for which the
identity holds to machine precision — is *not* the number Szigeti actually
computed. The impact is small but real:

```{r val-legacy}
do.call(rbind, lapply(names(sb), function(nm) {
  s <- cgr_strata(sb[[nm]]); m <- lapply(s, mean); o <- cgr_observed(s)
  ex <- cgr_ratios(s, legacy_round = FALSE)   # exact
  lg <- cgr_ratios(s, legacy_round = TRUE)    # Szigeti's round(., 2)
  data.frame(outcome = nm,
             unadj_exact  = round(cgr_delta(o, m, ex$r, ex$s), 3),
             unadj_legacy = round(cgr_delta(o, m, lg$r, lg$s), 3),
             adj_exact    = round(cgr_delta(0.5, m, ex$r, ex$s), 3),
             adj_legacy   = round(cgr_delta(0.5, m, lg$r, lg$s), 3),
             shift = round(cgr_delta(o, m, lg$r, lg$s) -
                           cgr_delta(o, m, ex$r, ex$s), 3))
}))
```

Rounding shifts the unadjusted PANAS estimate by about +0.01 and Energy VAS by
about +0.02 — negligible for any conclusion, but it is what Szigeti ran, so it
is the correct baseline when the goal is to match the published table exactly
rather than to state the exact estimand. The Bayesian sections keep the exact
ratios, because there the object of interest is the estimand itself, not a
byte-for-byte reproduction of the original script.

## The reference-line test

The identity above makes the position of the reference line *checkable* rather
than decorative: since $\Delta(c_{obs})$ equals the paper's own reported
unadjusted estimate, a line drawn at some claimed CGR is only in the right place
if $\Delta$ there equals that value. The paper's Figure 4 draws its "original
CGR" line at **0.72** (Fig 4 caption; see `reports/SOURCES.md`), so this tests
0.72 against the computed observed CGR.

```{r val-refline}
pub_unadj <- c(PANAS = 3.2, `Mood VAS` = 6.4, `Energy VAS` = 11.5,
               `Cognitive performance` = 0.0)
do.call(rbind, lapply(names(sb), function(nm) {
  z <- cgr_reference_line_test(sb[[nm]], orig_cgr = 0.72,
                               published_unadj = pub_unadj[[nm]])
  data.frame(outcome = nm,
             published_unadj = z$published_unadj,
             computed_obs_cgr = round(z$computed_obs_cgr, 4),
             D_at_obs = round(z$D_at_obs, 2),
             err_at_obs = round(z$err_at_obs, 2),
             D_at_0.72 = round(z$D_at_orig_cgr, 2),
             err_at_0.72 = round(z$err_at_orig, 2))
}))
```

Read `err_at_obs` against `err_at_0.72`. At the computed CGR of 0.647 the curve
lands on the paper's own reported unadjusted values to within ~0.12; at 0.72 it
overshoots them by 1.0–2.1 points on the three self-report scales. So the green
line as drawn is **not** the point where the curve equals the reported
unadjusted estimate — the signature of a misplaced reference line.

**The source code settles why.** In
[`szb37/CorrectGuessRateCurve`](https://github.com/szb37/CorrectGuessRateCurve)
the reference line is drawn from `trial_cgrs = {'sbmd': 0.72}` in
`src/config.py` — a **hardcoded constant**, not a quantity computed from the
data. The same code *does* compute the trial's CGR from the data as
`(n_plpl + n_acac) / n` (= 0.647), but that computed value is not what the plot's
green line uses. So 0.72 is a fixed annotation that does not match the data's
correct guess rate; it coincides almost exactly with the **placebo-arm**
correct-guess rate (0.7234, Section 10). This is no longer a hypothesis — see
`reports/UNRESOLVED.md` U3 and `reports/SOURCES.md`.

## Original KDE procedure vs the analytic curve

This is the section that separates the three possible sources of difference.

```{r val-kde, cache=TRUE}
set.seed(11)
# 32 is the resample count the author's code actually used for the CGR curve
# (config.py cgrC_low); 100 is the count the paper text describes.
kde_tab <- do.call(rbind, lapply(names(sb), function(nm) {
  d <- sb[[nm]]; o <- cgr_observed(cgr_strata(d))
  rbind(
    cbind(outcome = nm, target = "observed",
          cgr_kde_ladder(d, o,   reps = c(32, 100, 1000, 10000))),
    cbind(outcome = nm, target = "0.50",
          cgr_kde_ladder(d, 0.5, reps = c(32, 100, 1000, 10000))))
}))
kde_tab[, c("outcome","target","n_rep","kde_est","kde_mcse","mean_p",
            "analytic","diff_from_analytic")]

```

Three conclusions:

1. **KDE versus a Gaussian stratum model contributes essentially nothing.** At
   10 000 resamples the KDE average converges on the analytic value in every
   cell. The estimand uses only stratum means, and KDE smoothing preserves
   means.
2. **Few resamples is materially noisy.** At 32 resamples — the count the repo's
   Figure-4 config specifies (see the caveats in Section 2) — the Monte Carlo SE
   is roughly 0.2–0.5 points on the PANAS/VAS scales, about $1.8\times$ the
   100-resample value and larger than the 0.12 discrepancy this document flags
   in the paper's reference-line placement. PANAS at CGR 0.50 wanders by tens of
   a percent purely from resampling noise.
3. **The averaged p-values reproduce the published ones** (PANAS ~0.41 vs 0.43;
   Energy ~0.043 vs 0.04), confirming the port is faithful rather than merely
   plausible.

So the difference between the original and this document is **computational
stability plus a change of inferential summary** — not a change of estimand.

## Conjugate versus matched JAGS

The JAGS prior has been **corrected**. It previously used independent priors
`mu[j] ~ dnorm(0, 1e-6)` and `tau[j] ~ dgamma(1e-3, 1e-3)`, which is *not* the
Normal-Inverse-Gamma model, because NIG requires
$\text{Var}(\mu \mid \sigma^2) = \sigma^2/k_0$, i.e. precision $k_0 \tau$. The
corrected model is

```

tau[j] ~ dgamma(a0, b0)
mu[j]  ~ dnorm(m0, k0 * tau[j])

```

with the hyperparameters passed through the data list.

```{r val-jags, eval=has_jags, cache=TRUE}
chk <- cgr_check_backends(panas, GRID, n_draws = 40000,
                          jags_args = list(n_iter = 10000, n_chains = 4))
data.frame(quantity = names(chk), value = unlist(lapply(chk, as.character)))
```

```{r val-jags-note, echo=FALSE, results='asis'}
if (!has_jags) cat(
"> **NOT RUN.** rjags is unavailable, so the claim that the two backends target",
"an identical posterior is **unverified**. Until `cgr_check_backends()` passes,",
"treat them as comparable weak-prior models, not identical ones. If it fails,",
"delete the identity claim rather than weakening the prior to match.\n")

```

```{r jags-resolution, echo=FALSE}
# Single source of truth for the DECISION-1 / U1 status, reused verbatim in the
# Unresolved-questions section so the document stays self-consistent whether or
# not this render had JAGS.
jags_resolution <- if (has_jags) {
  sprintf(paste0(
    "**Resolved in this render.** `cgr_check_backends()` ran against JAGS: ",
    "posterior means agree to %.2g across the grid (max |z| = %.2f relative to ",
    "Monte Carlo error, max R-hat = %.4f, min ESS = %.0f), so the two backends ",
    "target the same posterior and Section 8's identity claim is supported. ",
    "Reproducing this line requires a JAGS install; a render without rjags falls ",
    "back to the NOT RUN note in Section 8."),
    chk$max_abs_diff_mean, chk$max_abs_z, chk$max_rhat, chk$min_ess)
} else {
  paste0(
    "**Unverified in this render.** No JAGS was available, so ",
    "`cgr_check_backends()` did not run and the identity claim is unsupported. ",
    "Install JAGS and re-render to resolve it; if it fails, delete the claim ",
    "rather than weakening the conjugate prior to force agreement.")
}
```

---

# Simulation study

A single simulated trial is an illustration, not validation. This section runs
**500 independent trials per scenario**.

True $\Delta(0.5)$ equals the direct treatment effect, because at perfect
blinding the guess distribution is identical in both arms, so the AEB term
contributes equally to each and cancels.

```{r sim-run, cache=TRUE}
op <- cgr_operating(n_trials = 500, n = 230, p_cg = 0.7,
                    noise = "all", n_draws = 4000, seed = 1)
op
```

Reference values obtained while preparing this document:

| DTE | AEB | true | adj bias | RMSE | coverage | P(post>0.95) | freq p<.05 |
|---|---|---|---|---|---|---|---|
| off | off | 0 | 0.05 | 0.59 | 0.938 | 0.066 | 0.056 |
| on | off | 3 | 0.05 | 1.09 | 0.936 | 0.892 | 0.876 |
| off | on | 0 | 0.02 | 1.09 | 0.940 | 0.056 | 0.826 |
| on | on | 3 | 0.05 | 1.33 | 0.960 | 0.704 | 0.992 |

- The adjusted estimator is **unbiased in all four scenarios** (|bias| ≤ 0.05).
- **Credible interval coverage is nominal** (0.936–0.960 vs 0.95). This is the
  first direct evidence that the intervals mean what they claim.
- The false-favourable rate is controlled (0.056, 0.066) where the unadjusted
  frequentist test fires at 0.826.
- **Adjustment costs power**: 0.892 → 0.704 in the partial-mediation row. A real
  cost, reported rather than buried.

These are operating characteristics of the *adjusted estimand* under the AEB
data-generating model. They are **not** evidence that the Bayesian
implementation outperforms the KDE implementation — Section 8 shows both target
the same quantity. A like-for-like comparison of the two procedures' operating
characteristics has not been run; see Section 14.

## The generative-model ambiguity, resolved

Equation 4 does not state whether the sd of the gated DTE/AEB terms applies to
everyone (`noise = "all"`) or only to the affected subgroup (`noise = "arm"`).
Two independent lines of evidence settle it:

| Evidence | published | `"all"` | `"arm"` |
|---|---|---|---|
| Hedges' g (DTE on, AEB off) | 0.40 | **0.4011** | 0.5023 |
| Table 1 sig. rate, DTE off/AEB off | 0.05 | **0.056** | 0.056 |
| Table 1 sig. rate, DTE on/AEB off | 0.86 | **0.876** | 0.970 |
| Table 1 sig. rate, DTE off/AEB on | 0.78 | **0.826** | 0.930 |
| Table 1 sig. rate, DTE on/AEB on | 0.99 | **0.992** | 0.998 |

`"all"` is now the documented default on evidence, not convenience.

**Caveat.** The author's source code was **not located** — the public repository
serves the data file but returns 404 for source paths. This is convergent
empirical evidence, not verification against the original implementation.
Section 9's figures are therefore a **reproduction consistent with the published
operating characteristics**, not a verified line-by-line reproduction.

---

# Empirical microdosing results

## Data audit

```{r audit}
audit <- do.call(rbind, lapply(names(sb), function(nm) {
  code <- SCALES[[nm]]
  all_rows <- sum(raw$test_name == code)
  w1 <- raw[raw$test_name == code & raw$tp == "w1s1", ]
  d  <- sb[[nm]]; s <- cgr_strata(d)
  data.frame(outcome = nm, total_records_all_tp = all_rows,
             w1s1_records = nrow(w1),
             excluded_other_tp = all_rows - nrow(w1),
             missing_value = sum(is.na(w1$value)),
             duplicated_id = sum(duplicated(w1$trial_id)),
             final_n = nrow(d),
             ACAC = length(s$ACAC), ACPL = length(s$ACPL),
             PLAC = length(s$PLAC), PLPL = length(s$PLPL),
             obs_cgr = round(cgr_observed(s), 4))
}))
audit

```

### The 0.72 versus 0.647 discrepancy

**The source code settles this: 0.72 was never computed from the data.** In
[`szb37/CorrectGuessRateCurve`](https://github.com/szb37/CorrectGuessRateCurve)
the Figure 4 reference line is drawn from `trial_cgrs = {'sbmd': 0.72}` in
`src/config.py` — a hardcoded constant. The same code computes the trial's CGR
from the data as `(n_plpl + n_acac) / n` = 0.647 and does not use it for the
line. So the 0.72 is a fixed annotation, not the data's correct guess rate; the
public week-1 data gives 0.647 and no pooling reaches 0.72 (w1 0.647, w2 0.679,
w3 0.620, w4 0.651; pooled 0.649).

Where did the *number* 0.72 come from? The most likely origin is the
placebo-arm correct-guess rate — splitting by **actual allocation**:

```{r cgr-by-arm}
do.call(rbind, lapply(names(sb), function(nm) {
  d <- sb[[nm]]
  do.call(rbind, lapply(c("AC", "PL"), function(a) {
    z <- d[d$condition == a, ]
    data.frame(outcome = nm, actual_arm = a, n = nrow(z),
               correct = sum(z$condition == z$guess),
               rate = round(mean(z$condition == z$guess), 4))
  }))
}))
```

For PANAS the placebo-arm correct-guess rate is **0.7234**, essentially the
hardcoded 0.72, while the microdose arm is 0.5275 — barely above chance. That
the constant equals the placebo-arm rate is a strong coincidence, and the likely
reason the value 0.72 was chosen — but *that* part is a hypothesis; the fact that
0.72 is hardcoded and not the data's overall CGR is settled by the code.

It matters substantively. The arms are unbalanced (141 placebo vs 91 microdose
in week 1) and guesses skew heavily toward "placebo". A single scalar CGR hides
that asymmetry, which the estimand encodes through $s$.

### The n = 232 versus n = 233 discrepancy

**Unresolved.** No scale at any timepoint has n = 233. Eight scales at w1s1 have
exactly 232; there are no duplicate `trial_id` values and no missing outcome
values. The gap is one record and cannot be reproduced from the public file.
Flagged for the author.

## Results

```{r empirical, cache=TRUE}
set.seed(2)
# Evaluate the posterior on a grid that includes each outcome's *exact* observed
# CGR, so the unadjusted row is read at c_obs (e.g. 0.6466) rather than snapped
# to the nearest 0.01 grid point (0.65), which would overstate it by ~0.05.
fits <- lapply(sb, function(d) {
  o <- cgr_observed(cgr_strata(d))
  cgr_conjugate(d, sort(unique(c(GRID, o))), n_draws = N_DRAWS, direction = 1)
})

do.call(rbind, lapply(names(sb), function(nm)
  cgr_summary_table(fits[[nm]], cgr_observed(cgr_strata(sb[[nm]])), nm)))

```

The `pct_attenuation` column (e.g. ~66% for PANAS) is a **ratio between two
estimates and depends on the likelihood**: under the Student-$t$ fit of
Section 11 the same PANAS attenuation reads ~57%, not 66%. Treat it as a
rough descriptor, not a fixed quantity; `pct_attenuation` is also suppressed
(shown as `NA`) whenever the unadjusted estimate is not distinguishable from
zero, because a ratio to a near-zero denominator is meaningless.

```{r empirical-plot, fig.height=6}
cur <- do.call(rbind, lapply(names(sb), function(nm) {
  z <- fits[[nm]]; z$outcome <- nm; z }))
cgr_plot(cur[cur$outcome == "PANAS", ],
         obs_cgr = cgr_observed(cgr_strata(panas)),
         title = "PANAS, week 1 acute")
```

Note what the lower panel is and is not. It is
$P(\Delta(c) > 0 \mid y)$ — the posterior probability that the reweighted
contrast is positive. It is **not a p-value**, there is no line at 0.05, and the
line at 0.95 is a descriptive marker, not a universal cutoff.

For outcomes where lower scores are better (QIDS, STAIT), pass
`direction = -1` so favourability is declared rather than assumed.

---

# Sensitivity analyses

```{r sens-prior, cache=TRUE}
set.seed(3)
priors <- list(
  "vague (default)"   = list(m0 = 0,  k0 = 1e-6, a0 = 1e-3, b0 = 1e-3),
  "weakly informative"= list(m0 = 0,  k0 = 0.01, a0 = 1,    b0 = 1),
  "informative at 0"  = list(m0 = 0,  k0 = 1,    a0 = 2,    b0 = 50),
  "unit-information"  = list(m0 = 15, k0 = 1,    a0 = 1,    b0 = 100)
)
do.call(rbind, lapply(names(priors), function(p) {
  z <- cgr_conjugate(panas, GRID, n_draws = 8000, prior = priors[[p]])
  a <- z[which.min(abs(z$cgr - 0.5)), ]
  data.frame(prior = p, adj_mean = round(a$est, 3),
             lo = round(a$lo, 3), hi = round(a$hi, 3),
             p_fav = round(a$p_fav, 3))
}))

```

```{r sens-ratios, cache=TRUE}
# EXTENSION, not reproduction: the original estimand CONDITIONS on the observed
# within-class ratios. Treating them as uncertain is a different, larger
# uncertainty statement.
set.seed(4)
n_draws <- 8000
mu_d <- lapply(st, nig_draws, n_draws = n_draws)
rd <- cgr_ratio_draws(st, n_draws)

fixed  <- cgr_delta(0.5, mu_d, rat$r, rat$s)
random <- {
  w_acac <- 0.5 * (1 - rd$r); w_acpl <- 0.5 * rd$s
  w_plpl <- 0.5 * rd$r;       w_plac <- 0.5 * (1 - rd$s)
  (w_acac * mu_d$ACAC + w_acpl * mu_d$ACPL) / (w_acac + w_acpl) -
    (w_plpl * mu_d$PLPL + w_plac * mu_d$PLAC) / (w_plpl + w_plac)
}
data.frame(
  ratios = c("fixed (original estimand)", "Beta-uncertain (extension)"),
  mean = round(c(mean(fixed), mean(random)), 3),
  sd   = round(c(sd(fixed), sd(random)), 3),
  lo   = round(c(quantile(fixed, .025), quantile(random, .025)), 3),
  hi   = round(c(quantile(fixed, .975), quantile(random, .975)), 3),
  p_fav = round(c(mean(fixed > 0), mean(random > 0)), 3))
```

```{r sens-small, cache=TRUE}
# Small strata and skew: subsample to stress the normal within-stratum model.
set.seed(5)
do.call(rbind, lapply(c(1.0, 0.5, 0.25, 0.12), function(f) {
  idx <- sample.int(nrow(panas), max(40, round(nrow(panas) * f)))
  d <- panas[idx, ]
  ok <- tryCatch({
    z <- cgr_conjugate(d, GRID, n_draws = 6000)
    a <- z[which.min(abs(z$cgr - 0.5)), ]
    data.frame(fraction = f, n = nrow(d),
               min_stratum = min(lengths(cgr_strata(d))),
               adj = round(a$est, 3), width = round(a$hi - a$lo, 3),
               p_fav = round(a$p_fav, 3))
  }, error = function(e) data.frame(fraction = f, n = nrow(d),
      min_stratum = NA, adj = NA, width = NA, p_fav = NA))
  ok
}))

```

```{r sens-skew}
# Are the strata plausibly normal? Skewness and excess kurtosis per stratum.
do.call(rbind, lapply(names(sb), function(nm) {
  s <- cgr_strata(sb[[nm]])
  do.call(rbind, lapply(STRATA, function(k) {
    x <- s[[k]]; z <- (x - mean(x)) / sd(x)
    data.frame(outcome = nm, stratum = k, n = length(x),
               skew = round(mean(z^3), 3),
               excess_kurtosis = round(mean(z^4) - 3, 3),
               sd = round(sd(x), 2))
  }))
}))
```

Unequal stratum variances are handled by construction — each stratum has its own
$\sigma_j^2$, so no homoscedasticity assumption is imposed across strata. But the
skew table is not innocuous: PANAS stratum ACAC has skew $-0.94$ and excess
kurtosis $1.42$, which is exactly the regime where a Gaussian likelihood's
*intervals* (not its means) can be optimistic. So the robust check is worth
running rather than deferring.

A Student-$t$ likelihood with estimated degrees of freedom $\nu$ down-weights
outliers and has no conjugate form, so it is reachable only through JAGS.

```{r sens-student-t, eval=has_jags, cache=TRUE}
set.seed(6)
jn <- cgr_jags(panas, likelihood = "normal", n_iter = 10000, n_chains = 4)
jt <- cgr_jags(panas, likelihood = "t",      n_iter = 10000, n_chains = 4)
i5 <- function(x) which.min(abs(x$cgr - 0.5))
data.frame(
  likelihood = c("normal", "Student-t"),
  adj_at_0.5 = round(c(jn$est[i5(jn)], jt$est[i5(jt)]), 3),
  cri_95 = c(sprintf("[%.2f, %.2f]", jn$lo[i5(jn)], jn$hi[i5(jn)]),
             sprintf("[%.2f, %.2f]", jt$lo[i5(jt)], jt$hi[i5(jt)])),
  P_positive = round(c(jn$p_fav[i5(jn)], jt$p_fav[i5(jt)]), 3),
  est_df_nu = c(NA, round(attr(jt, "nu"), 1)))
```

```{r sens-student-t-note, echo=FALSE, results='asis'}
if (!has_jags) cat(
"> **NOT RUN.** rjags is unavailable, so the Student-*t* robustness check did",
"not execute. Install JAGS and re-render to run it.\n")
```

A **small** $\nu$ (say under 10) would mean the data are genuinely heavier-tailed
than Gaussian and the robust fit is doing real work; a **large** $\nu$ means the
$t$ collapsed back toward the normal. Here $\nu \approx 18$, so the tails are
only mildly heavy.

**But read the point estimate before concluding "no change".** The adjusted
estimate moves from 1.077 (normal) to ~1.355 (Student-$t$) — a **~26% shift** in
the headline number. That is not negligible for the *attenuation* summary
reported elsewhere: under the normal likelihood PANAS attenuates
$3.15 \to 1.08 \approx 66\%$, but under the Student-$t$ it is
$3.15 \to 1.355 \approx 57\%$. So the percentage attenuation is
**likelihood-dependent** and should not be quoted as a single fixed number.

What *does* survive the switch is the **inference**: the two posteriors differ by
only ~0.2 of a posterior SD, their 95% credible intervals overlap heavily, and
$P(\text{effect} > 0)$ moves only from 0.79 to 0.84. So the qualitative
conclusion — a positive effect that is much attenuated but not clearly zero —
holds under both likelihoods; it is the exact attenuation *percentage* that is
not robust, not the direction.

---

# A region of practical equivalence

The lower panel of Section 10 reports $P(\Delta(c) > 0 \mid y)$. As an
inferential summary a direction probability — like the two-sided tail it
replaced — has two defects. It is **magnitude-blind**: a tiny, precisely
estimated effect and a large, uncertain one can score identically. And it
**conflates two states of knowledge**: a probability near 0.5 can mean "the
posterior is concentrated near zero, so the effect is negligible" or "the
posterior is diffuse, so we have learned nothing" — opposite conclusions the
statistic cannot separate. For a method whose whole purpose is asking whether an
apparent effect *survives* adjustment, that is the distinction that matters.

The fix is not a Bayes factor against $\Delta = 0$: with the vague default prior
($k_0 = 10^{-6}$) it depends on the prior width without limit — the
Jeffreys–Lindley paradox — so it would measure the prior, not the data. Instead
declare a band $[-\delta, +\delta]$ of practically-negligible effects and report
the three exhaustive probabilities

$$P(\Delta < -\delta), \qquad P(|\Delta| \le \delta), \qquad P(\Delta > +\delta),$$

which sum to 1 and whose middle term separates "negligible" from "uninformative".
The cost — $\delta$ must be declared — is a feature: it forces the
practical-significance question into the open. The default is
$\delta = 0.1 \times$ the outcome's pooled SD.

```{r rope-run, cache=TRUE}
set.seed(31)
rope <- lapply(sb, cgr_rope, n_draws = N_DRAWS, direction = 1)
do.call(rbind, lapply(names(sb), function(nm) {
  z <- rope[[nm]]; o <- cgr_observed(cgr_strata(sb[[nm]]))
  do.call(rbind, lapply(list(c(o, 0), c(0.5, 1)), function(tt) {
    a <- z[which.min(abs(z$cgr - tt[1])), ]
    data.frame(outcome = nm,
               at = if (tt[2] == 0) "observed CGR" else "CGR 0.50",
               ROPE_delta = round(a$delta, 2),
               post_mean = round(a$est, 3),
               P_harm = round(a$p_harm, 3),
               P_negligible = round(a$p_negligible, 3),
               P_benefit = round(a$p_benefit, 3))
  }))
}))
```

Read the `CGR 0.50` rows. For **PANAS**, adjustment leaves roughly a 40% chance
the effect is practically negligible against roughly a 50% chance it is
meaningfully positive — genuine uncertainty, not a demonstrated null; reporting
"not significant" there would be a misreading. For **Energy VAS**, the
probability of a meaningful benefit stays high even after adjustment. For
**cognitive performance**, the mass is spread across all three regions — the
honest description of a small study with a small ROPE: uninformative rather than
null.

```{r rope-plot, cache=TRUE, fig.height=4.2}
stack <- do.call(rbind, lapply(names(sb), function(nm) {
  z <- rope[[nm]]
  do.call(rbind, lapply(c("p_benefit", "p_negligible", "p_harm"), function(k)
    data.frame(cgr = z$cgr, outcome = nm, p = z[[k]], region = k,
               obs = cgr_observed(cgr_strata(sb[[nm]])))))
}))
stack$outcome <- factor(stack$outcome, levels = names(SCALES))
stack$region  <- factor(stack$region,
  levels = c("p_benefit", "p_negligible", "p_harm"),
  labels = c("meaningful benefit", "practically negligible", "meaningful harm"))

ggplot(stack, aes(cgr, p, fill = region)) +
  geom_area(position = "stack") +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "black") +
  geom_vline(aes(xintercept = obs), linetype = "dashed", colour = "darkgreen") +
  scale_fill_manual(values = c("meaningful benefit" = "#2471A3",
                               "practically negligible" = "grey75",
                               "meaningful harm" = "#C0392B")) +
  scale_y_continuous(expand = c(0, 0)) +
  facet_wrap(~ outcome, nrow = 1) +
  labs(x = "Correct guess rate (CGR)", y = "posterior probability", fill = NULL,
       title = "ROPE decomposition of the CGR-adjusted effect",
       subtitle = paste("regions are exhaustive and sum to 1 at every CGR;",
                        "black = perfect blinding, green = observed CGR")) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

**Sensitivity to the width.** A region-based conclusion is only as robust as the
region, so vary $\delta$ from 0.05 to 0.30 SD at CGR 0.50:

```{r rope-sens, cache=TRUE}
set.seed(41)
do.call(rbind, lapply(names(sb), function(nm) {
  s <- cgr_rope_sensitivity(sb[[nm]], at_cgr = 0.5, n_draws = 12000)
  data.frame(outcome = nm, delta_in_SD = s$delta_in_SD,
             delta = round(s$delta, 2),
             P_negligible = round(s$p_negligible, 3),
             P_benefit = round(s$p_benefit, 3))
}))
```

Energy VAS is the only outcome whose conclusion is stable across every width —
the probability of a meaningful benefit stays high even at $\delta = 0.3$ SD.
PANAS and Mood VAS flip from "probably meaningful" to "probably negligible" as
$\delta$ widens, exactly the fragility a single tail number would have concealed.

---

# Assumptions and limitations

**This is a sensitivity analysis. It does not identify a direct pharmacological
effect.**

1. **Malicious versus benign unblinding.** The method assumes unblinding is
   *malicious*: side effects reveal allocation, and expectancy then drives
   outcome (PT → TE → OUT). If unblinding is *benign* — people guess correctly
   *because* they improved — the causal arrow runs OUT → PT, and CGR adjustment
   removes a genuine treatment effect. **The original paper warns about this
   explicitly and it must be preserved prominently.** The authors argue for
   malicious unblinding here on the grounds that 55% cited bodily sensations
   versus 23% citing mental benefits, and that the placebo-microdose difference
   is small relative to natural day-to-day variability. That is an argument from
   evidence, not a demonstration.
2. **Collider bias.** Guess is a downstream consequence of both treatment and
   outcome. Conditioning on it can induce association where none exists.
3. **Transportability.** The four observed strata are assumed to represent the
   corresponding strata in a hypothetical perfectly blinded population. People
   who guess correctly under poor blinding may not be exchangeable with those
   who would guess correctly by chance under good blinding.
4. **Conditioning on observed guess behaviour.** The estimand holds $r$ and $s$
   fixed at their observed values, treating stratum composition as known rather
   than estimated (Section 11 quantifies the alternative).
5. **Normality within stratum.** Bounded scales with floor/ceiling effects are
   not exactly normal (Section 11 reports skew and kurtosis).
6. **Small strata.** ACPL and PLAC hold 39–43 observations for PANAS. Cognitive
   scales have fewer still.
7. **Binary guess.** Treatment belief is reduced to one bit, with no confidence
   rating, so a confident correct guess and a lucky one are indistinguishable.
8. **Association is not causation.** Removing an association between outcome and
   guessing is not the same as removing expectancy causally.
9. **Attenuation is not an effect decomposition.** The gap between
   $\Delta(c_{obs})$ and $\Delta(0.5)$ is *not* a measurement of "how much was
   expectancy". It is the change in the estimand under a reweighting assumption.
10. **Reproducing the curve does not validate the assumptions.** It validates
    the arithmetic.

---

# What the Bayesian approach adds

Stated narrowly, because Section 8 showed both approaches target the same
estimand.

**What it adds:**

- **A quantity with an interpretation.** $P(\Delta(0.5) > 0 \mid y)$ is a
  probability statement about the parameter under the model. An average of 100
  p-values is not a p-value for anything.
- **Removal of an artefactual error term.** The 0.12–0.29 point Monte Carlo SE
  from 100 resamples is a property of the computation, not the trial. The
  conjugate posterior is exact.
- **Direct uncertainty propagation.** The credible interval comes from the same
  object as the point estimate, rather than being assembled from resampled
  fits.
- **Extensibility.** Robust likelihoods, uncertain ratios, and hierarchical
  structure are reachable in the same framework.
- **Calibration that has been checked.** Section 9 measured coverage rather than
  assuming it.

**What it does not add:**

- No extra information about participants. `n_draws` is not a sample size.
- No relaxation of any causal assumption in Section 12.
- No demonstrated superiority over the KDE procedure as an estimator, since
  both converge to the same value.

---

# Unresolved questions

1. **Corrected JAGS prior.** `r jags_resolution`
2. **n = 232 versus the published n = 233.** The source paper (Szigeti 2023)
   states n = 233 week-1 datapoints under exactly the `w1s1` filter used here, so
   the analysis rule is confirmed; the public file still gives 232. A genuine
   one-record gap, unexplained by the papers. See `reports/SOURCES.md`.
3. **The 0.72 CGR — explained.** The paper's Fig 4 caption states the trial CGR
   is 0.72, but that equals the *placebo-arm* correct-guess rate (0.7234); the
   analyzed data's overall CGR is 0.647 (which reproduces the paper's Table 2),
   and the author's 2024 review cites ~65–70%. The Fig 4 reference line is
   therefore misplaced by ~0.07 — the error the Section 4 identity check catches.
   Worth one confirmation from the author. See `reports/SOURCES.md`.
4. **Exact AEB simulation source not located.** The `noise = "all"` resolution
   now matches the paper's published Table 1 (0.05/0.86/0.78/0.99) and Hedges
   g = 0.4, but the exact Eq. 4 SD scope lives in the paper's Supplementary
   Table 1, so this is not yet a line-by-line code check.
5. **No head-to-head operating-characteristic comparison** of the KDE procedure
   against the Bayesian one. Section 9 characterises the estimand, not the two
   procedures against each other.
6. **Benign versus malicious unblinding is not testable** with these data.
7. **KDE bandwidth.** The original uses sklearn's fixed default of 1.0. Because
   the estimand uses only means this does not matter here, but it was never a
   deliberate choice and would matter for any tail-based estimand.

---

# Sources

Full per-finding citations are in `reports/SOURCES.md`.

- **Method paper (reproduced here).** Szigeti B, Nutt D, Carhart-Harris R,
  Erritzoe D. "The difference between 'placebo group' and 'placebo control': a
  case study in psychedelic microdosing." *Scientific Reports* 13:12107 (2023).
  <https://doi.org/10.1038/s41598-023-34938-7>
- Szigeti B, Weiss B, Rosas FE, Erritzoe D, Nutt D, Carhart-Harris R. "Assessing
  expectancy and suggestibility in a trial of escitalopram v. psilocybin for
  depression." *Psychological Medicine* 54, 1717–1724 (2024).
  <https://doi.org/10.1017/S0033291723003653>
- Szigeti B, Heifets BD. "Expectancy Effects in Psychedelic Trials."
  *Biological Psychiatry: Cognitive Neuroscience and Neuroimaging* 9:512–521
  (2024). <https://doi.org/10.1016/j.bpsc.2024.02.004>
- Public dataset: Szigeti B, et al. "Self-blinding citizen science to explore
  psychedelic microdosing." *eLife* 10:e62878 (2021).
  <https://doi.org/10.7554/eLife.62878>

---

# Session information and reproducibility

```{r session}
cat("R version:      ", R.version.string, "\n")
cat("Platform:       ", R.version$platform, "\n")
cat("Seed:            20260721 (set in setup chunk)\n")
cat("Data SHA-256:    86aa784528ee045c61fadf3eacfd3e1897d16aae9839cee7cb4bfe839a7cc4e3\n")
cat("Data accessed:   2026-07-21\n")
cat("JAGS available: ", has_jags, "\n")
if (has_jags) cat("JAGS version:   ", as.character(rjags::jags.version()), "\n")
sessionInfo()
```
