---
title: "The Correct Guess Rate Curve as a usable analysis tool"
subtitle: "A Bayesian reimplementation, an UNKNOWN-guess extension, simulation-based design checks, and an interactive app"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r s-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
set.seed(20260724)
options(width = 100)

# This manuscript consumes the installed cgrc.bayes package; every cgr_* /
# cgrc*/ nig_draws() call below is the package's implementation. The exhaustive
# technical record - full derivations, code audit, extended diagnostics, prior
# and robustness sensitivities, ROPE analyses and the complete external-data
# provenance - is the companion document CGRC_bayes.Rmd. This summary is the
# focused story.
library(cgrc.bayes)
library(ggplot2)

# --- primary data: the public week-1 microdose trial ------------------------
# Put pacutes.csv in the working directory (fetch + checksum-verify with
# data-raw/download_data.R). MD = microdose/active, PL = placebo.
stopifnot(file.exists("pacutes.csv"))
raw <- read.csv("pacutes.csv", stringsAsFactors = FALSE)
load_scale <- function(code, tp = "w1s1") {
  d <- raw[raw$test_name == code & raw$tp == tp, ]
  data.frame(condition = ifelse(d$condition == "MD", "AC", "PL"),
             guess     = ifelse(d$guess     == "MD", "AC", "PL"),
             value     = d$value, stringsAsFactors = FALSE)
}
SCALES <- c("PANAS" = "PANAS", "Mood VAS" = "mood",
            "Energy VAS" = "energy", "Cognitive performance" = "CPS")
sb    <- lapply(SCALES, load_scale); names(sb) <- names(SCALES)
panas <- sb[["PANAS"]]

# --- external trial datasets ------------------------------------------------
# Put these CSVs in the working directory alongside pacutes.csv.
read_ext <- function(f) {
  stopifnot(file.exists(f))
  utils::read.csv(f, stringsAsFactors = FALSE)
}

# --- the app's "Analyse your own trial" logic, as a helper ------------------
# Audit inputs, detect binary vs UNKNOWN guesses, then run the four-stratum
# cgrc() or the six-stratum cgrc_unknown(). This is the same decision the Shiny
# app makes; reused here so the worked examples are literally the app workflow.
app_analyse <- function(df, condc, guessc, valc, direction = 1, delta = NULL,
                        delta_sd_frac = 0.5, n_draws = 8000) {
  aud   <- cgrc_input_audit(df[[condc]], df[[guessc]], df[[valc]])
  clean <- aud$clean
  mode  <- if (aud$has_unknown) "unknown" else "binary"
  if (mode == "binary") clean <- clean[clean$guess %in% c("AC", "PL"), , drop = FALSE]
  d <- if (!is.null(delta)) delta else delta_sd_frac * stats::sd(clean$value)
  fit <- if (mode == "unknown")
           cgrc_unknown(clean, n_draws = n_draws, direction = direction, seed = 1)
         else cgrc(clean, n_draws = n_draws, direction = direction, seed = 1)
  hd  <- if (mode == "unknown")
           cgrc_unknown_headline(clean, direction = direction, delta = d, n_draws = n_draws, seed = 1)
         else cgrc_headline(clean, direction = direction, delta = d, n_draws = n_draws, seed = 1)
  list(mode = mode, audit = aud, trial = clean, dir = direction, delta = d, fit = fit, head = hd)
}
app_curve <- function(a, label) {
  if (a$mode == "unknown")
    cgr_unknown_plot(a$fit$curve, obs_cgr = a$fit$observed_directional_cgr,
                     u = a$fit$target_unknown_rate, direction_label = label)
  else cgr_plot(a$fit$curve, obs_cgr = a$fit$observed_cgr, direction_label = label)
}

N_DRAWS  <- 20000
has_jags <- requireNamespace("rjags", quietly = TRUE) &&
            requireNamespace("coda",  quietly = TRUE)
```

# Motivation and contribution

Blinding fails in many trials, and psychedelic and microdose trials fail it
badly: participants who correctly guess their arm can respond partly to *belief*
rather than drug, and because more people in the active arm guess correctly, that
belief-driven response accumulates unevenly on the treatment side. The **Correct
Guess Rate Curve (CGRC)** of Szigeti et al. (2023) asks a counterfactual: *what
would the estimated effect have been if the correct-guess rate (CGR) had been 50%
— i.e. if blinding had held?* It reweights the four treatment × guess strata to a
target guess rate while holding the within-class arm composition fixed. The
original implementation obtains that estimate — and its uncertainty — by resampling
pseudo-datasets from kernel-density estimates of the strata and **averaging the
resulting estimates and p-values** over resamples.

`cgrc.bayes` keeps the same CGRC estimand but replaces the resampling with a
**posterior**: a Normal-Inverse-Gamma model for the four stratum means whose
uncertainty propagates analytically through the reweighting. Over time the project
grew from that reimplementation into a small, usable CGRC workflow.

> This work reproduces the original CGRC, formulates the same estimand in a
> Bayesian framework, evaluates its operating characteristics by simulation,
> extends it to trials allowing an "I do not know" treatment guess, and packages
> the resulting workflow in an interactive application exercised on several
> independent trial datasets.

Those are not five separate contributions; they are the components a researcher
needs to actually *use* CGRC on a real trial: a coherent estimate with honest
uncertainty, a way to check whether the adjustment is trustworthy for a given
design, a way to handle real blinding questionnaires that permit "I do not know",
and an interface that exposes all of it without custom code.

One caution frames everything below. **Attenuation is not a causal estimate.** When
the PANAS effect shrinks from a raw ~3.2 points to ~1 point at perfect blinding,
that drop is *not* "the fraction of the treatment effect caused by expectancy". The
CGRC is a sensitivity analysis for unblinding under an explicit reweighting model,
not a causal decomposition.

# Original CGRC and the estimand

Each participant contributes an arm (active AC / placebo PL) and a guess (AC / PL),
giving four strata `ACAC`, `ACPL`, `PLAC`, `PLPL` (arm, then guess). With $\bar y_k$
the stratum mean and $\rho_k$ its sample share, the **observed correct-guess rate**
is $c_{obs} = \rho_{ACAC} + \rho_{PLPL}$, and two within-class ratios describe how
correct- and incorrect-guessers split across arms:
$$
r=\frac{\rho_{PLPL}}{\rho_{PLPL}+\rho_{ACAC}},\qquad
s=\frac{\rho_{ACPL}}{\rho_{ACPL}+\rho_{PLAC}}.
$$
At a target guess rate $c$ the strata receive weights
$w_{ACAC}=c(1-r)$, $w_{PLPL}=cr$, $w_{ACPL}=(1-c)s$, $w_{PLAC}=(1-c)(1-s)$, and the
CGR-adjusted contrast is the reweighted active-minus-placebo difference
$$
\Delta(c)=\frac{w_{ACAC}\bar y_{ACAC}+w_{ACPL}\bar y_{ACPL}}{w_{ACAC}+w_{ACPL}}
        -\frac{w_{PLPL}\bar y_{PLPL}+w_{PLAC}\bar y_{PLAC}}{w_{PLPL}+w_{PLAC}}.
$$
The target of interest is $\Delta(0.5)$, the effect at a 50% guess rate. Because $r$
and $s$ come from the observed composition, at $c=c_{obs}$ the reweighting is a
no-op and
$$
\Delta(c_{obs})=\bar y_{AC}-\bar y_{PL},
$$
the ordinary unadjusted arm-mean difference. This **observed-CGR identity** is exact
and is the cleanest validation check in the project: a curve is only in the right
place if it passes through the reported unadjusted effect at the observed CGR.

```{r s-estimand}
st  <- cgr_strata(panas); rat <- cgr_ratios(st); o <- cgr_observed(st)
idn <- cgr_reference_line_test(panas, orig_cgr = o)
knitr::kable(
  data.frame(stratum = STRATA, n = lengths(st)[STRATA],
             mean = round(vapply(st[STRATA], mean, numeric(1)), 2)),
  caption = sprintf("PANAS week 1: four strata; observed CGR = %.4f, r = %.3f, s = %.3f",
                    o, rat$r, rat$s))
```

For PANAS the identity holds to machine precision: $\Delta(c_{obs})=$
`r round(idn$D_at_obs, 3)` equals the raw arm-mean difference
`r round(idn$raw_mean_diff, 3)` (error
`r sprintf("%.1e", abs(idn$D_at_obs - idn$raw_mean_diff))`), while
$\Delta(0.5)=$ `r round(cgr_delta(0.5, lapply(st, mean), rat$r, rat$s), 3)`. That
reweighted ~1-point value at perfect blinding, against the ~3.2-point raw
difference, is the CGRC's substantive claim for PANAS. (The reproduction of
Szigeti's *shipped* code uses a rounded-ratio variant, `legacy_round = TRUE`; the
estimand itself uses the exact ratios above. The two paths are kept distinct and
compared in the full Rmd.)

# Bayesian implementation of the same estimand

Each stratum is modelled as $y_{ij}\mid\mu_j,\sigma^2_j\sim N(\mu_j,\sigma^2_j)$ with
a Normal-Inverse-Gamma prior
$\mu_j\mid\sigma^2_j\sim N(m_0,\sigma^2_j/k_0)$,
$\sigma^2_j\sim\text{Inv-Gamma}(a_0,b_0)$, whose hyperparameters have plain meanings
($m_0$ a prior mean, $k_0$ a prior sample size in pseudo-observations, $(a_0,b_0)$
the variance shape and scale). With vague defaults ($m_0=0$, $k_0=10^{-6}$,
$a_0=b_0=10^{-3}$) the prior contributes negligibly against dozens of observations;
the conjugate posterior is available in closed form.

> For every posterior draw of the four stratum means, evaluate the same CGRC
> contrast $\Delta(c)$. The result is a posterior distribution over the whole CGRC
> curve — a posterior mean, a 95% credible interval, and the posterior probability
> the effect is favourable at each guess rate — rather than a collection of
> pseudo-dataset estimates and averaged p-values.

The number of draws controls only Monte Carlo error in these summaries; it is not a
sample size. (The draw order, within-draw dependence and full derivation are in the
companion Rmd.)

```{r s-bayes}
fit_panas <- cgrc(panas, n_draws = N_DRAWS, seed = 2)
knitr::kable(fit_panas$summary[, c("what", "post_mean", "cri_lo", "cri_hi", "p_favourable")],
             digits = 3, caption = "PANAS: CGR-adjusted posterior (Normal-Inverse-Gamma conjugate)")
```

Two probabilities carry the practical message and are **not** the same quantity:
$P(\text{favourable})$ — the posterior probability the adjusted effect is on the
beneficial side of zero — answers "is there an effect", while $P(\text{meaningful})$
— the probability it exceeds a pre-declared minimally-important difference — answers
"is it large enough to matter". A tail probability is a direct statement about the
parameter given the data, not a p-value.

# Reproduction and validation against Szigeti et al.

Four checks establish that the Bayesian version reproduces the original behaviour
and does not redefine the target.

**(1) Published behaviour is reproduced.** Under the activated-expectancy-bias (AEB)
generative model at the paper's regime ($n=230$, CGR $0.7$, a 3-point effect), the
unadjusted two-sided significance rates across the four (effect, expectancy)
scenarios reproduce the published Table 1 (0.05 / 0.86 / 0.78 / 0.99). This same
500-trial run is the operating-characteristics simulation used in Section 6.

```{r s-repro, cache=TRUE}
op <- cgr_operating(n = 230, p_cg = 0.7, mu_dte = 3, mu_aeb = 7.7,
                    n_trials = 500, seed = 1)
scen <- c("no effect / no expectancy", "effect / no expectancy",
          "no effect / expectancy", "effect / expectancy")
knitr::kable(data.frame(scenario = scen,
  `unadjusted p<0.05 (2-sided)` = round(op$freq_sig, 3),
  `published Table 1` = c(0.05, 0.86, 0.78, 0.99), check.names = FALSE),
  caption = "AEB simulation, n = 230, CGR = 0.7, 500 trials: reproduces the published significance rates")
```

**(2) The observed-CGR identity holds** for every outcome — the adjusted curve
recovers the ordinary unadjusted contrast at the observed CGR, to machine precision.

```{r s-identity}
knitr::kable(do.call(rbind, lapply(names(sb), function(nm) {
  z <- cgr_reference_line_test(sb[[nm]], orig_cgr = cgr_observed(cgr_strata(sb[[nm]])))
  data.frame(outcome = nm, observed_CGR = round(z$computed_obs_cgr, 4),
             `Delta(c_obs)` = round(z$D_at_obs, 3), `raw AC-PL` = round(z$raw_mean_diff, 3),
             `abs error` = sprintf("%.1e", abs(z$D_at_obs - z$raw_mean_diff)),
             check.names = FALSE)
})), caption = "Observed-CGR identity across the four outcomes")
```

**(3) KDE resampling converges on the analytic $\Delta(c)$.** As the resample count
grows the KDE average converges on the analytic value; the only systematic
difference at small counts is Monte Carlo variation. Because $\Delta(c)$ depends on
the strata only through their means and kernel smoothing preserves means, the KDE
step does not materially change the estimand.

```{r s-kde, cache=TRUE}
set.seed(11)
kl <- cgr_kde_ladder(panas, cgr_observed(cgr_strata(panas)), reps = c(32, 100, 1000, 10000))
knitr::kable(kl[, c("n_rep", "kde_est", "kde_mcse", "analytic", "diff_from_analytic")],
             digits = 3, caption = "KDE resampling converges on the analytic Delta (PANAS, observed CGR)")
```

**(4) The procedures agree on the estimand but summarise uncertainty differently.**
Over repeated trials the KDE point estimate and the Bayesian posterior mean of
$\Delta(0.5)$ are almost perfectly correlated with the same small bias; their
*decisions* differ, because averaging a p-value across resampled pseudo-datasets
has no accepted sampling interpretation.

```{r s-headtohead, cache=TRUE}
set.seed(101); N <- 300; n_i <- 200; p_cg <- 0.7; mu_dte <- 3
ke <- kp <- be <- bp <- rep(NA_real_, N)
for (i in seq_len(N)) {
  di <- sim_aeb(n_i, p_cg, dte_on = TRUE, aeb_on = FALSE, mu_dte = mu_dte)
  if (!all(STRATA %in% paste0(di$condition, di$guess))) next
  ki <- cgr_kde(di, 0.5, n_rep = 100); ke[i] <- ki$est; kp[i] <- ki$p
  sti <- cgr_strata(di); rti <- cgr_ratios(sti)
  ddi <- cgr_delta(0.5, lapply(sti, nig_draws, n_draws = 4000), rti$r, rti$s)
  be[i] <- mean(ddi); bp[i] <- mean(ddi > 0)
}
ok <- !is.na(kp)
knitr::kable(data.frame(
  Quantity = c("Point-estimate correlation (KDE vs posterior)",
               "Point-estimate bias: KDE / posterior",
               "Flag rate: KDE resample-averaged two-sided p < 0.05",
               "Flag rate: posterior P(favourable) > 0.975 (matches two-sided 0.05)",
               "Trials where the two decisions agree"),
  Value = c(sprintf("%.3f", cor(ke[ok], be[ok])),
            sprintf("%+.3f / %+.3f", mean(ke[ok]) - mu_dte, mean(be[ok]) - mu_dte),
            sprintf("%.3f", mean(kp[ok] < 0.05)),
            sprintf("%.3f", mean(bp[ok] > 0.975)),
            sprintf("%.0f%%", 100 * mean((kp[ok] < 0.05) == (bp[ok] > 0.975)))),
  check.names = FALSE),
  caption = "Original vs Bayesian inference, 300 simulated trials (true Delta = 3)")
```

The procedures agree closely on the CGRC point estimand but summarise uncertainty
differently; retaining posterior draws yields directly interpretable quantities in
place of an averaged p-value. This is a difference in inferential summary, not a
claim that either approach is categorically better.

The posterior flag here is thresholded at $P(\text{favourable})>0.975$ *only* to
match Szigeti's original **two-sided** $p<0.05$ decision on equal footing (a two-sided
5% test splits into 2.5% per tail, so the matching one-sided posterior bar is
$1-0.025$). This is a head-to-head decision-agreement check, not a reporting rule;
everywhere else in this document, and in the app, the reporting threshold is the
standard $P(\text{favourable})>0.95$.

# Empirical microdosing example

The primary worked example is the trial CGRC was built on. Following the paper, only
week-1, single-timepoint records are used — one observation per participant.

The public week-1 data yield an overall CGR of 0.647, whereas the published Figure 4
marks 0.72. The available plotting code uses 0.72 as a fixed annotation; the
observed-CGR identity places the unadjusted contrast at 0.647. This is noted here
primarily because the identity provides a useful implementation check.

```{r s-empirical}
emp <- do.call(rbind, lapply(names(sb), function(nm) {
  d <- sb[[nm]]; f <- cgrc(d, n_draws = N_DRAWS, seed = 2)
  h <- cgrc_headline(d, n_draws = N_DRAWS, seed = 2); s <- f$summary
  data.frame(outcome = nm, n = nrow(d), observed_CGR = round(f$observed_cgr, 3),
             `raw` = round(s$post_mean[1], 2), `adjusted (0.50)` = round(s$post_mean[2], 2),
             `95% CrI` = sprintf("[%.2f, %.2f]", s$cri_lo[2], s$cri_hi[2]),
             `P(fav)` = sprintf("%.0f%%", 100 * s$p_favourable[2]),
             `P(meaningful)` = sprintf("%.0f%%", 100 * h$p_meaningful_blind),
             check.names = FALSE)
}))
knitr::kable(emp, caption = "CGR-adjusted results, four outcomes (adjusted = perfect blinding, CGR 0.50; meaningful = beyond a half-SD difference)")
```

```{r s-panas-fig, fig.height=6}
plot(fit_panas, title = "PANAS, week 1")
```

The table reproduces the published attenuation pattern: PANAS and the mood/energy
VAS effects shrink substantially at a 50% guess rate, with cognitive performance
near zero throughout.

# Simulation study: is CGRC trustworthy for *this* design?

Szigeti et al. explicitly recommended that researchers first simulate the
performance of CGR adjustment for the parameters of their own application.
`cgr_operating()` operationalises exactly that recommendation: it runs many
independent trials under the AEB generative model at a chosen design and returns the
adjustment's operating characteristics. This is the section that turns CGRC from an
equation into something you can decide whether to trust.

## Main operating-characteristic simulation

The 500-trial run from Section 4 (`op`) spans the four scenarios: no effect, a real
direct treatment effect (DTE), an expectancy confound (AEB), and both. The
"unadjusted flag" is the posterior probability of a favourable effect from an
*unadjusted* arm-contrast; the "adjusted flag" is the same probability from the
CGR-adjusted estimate — both at the standard $P>0.95$.

```{r s-sim-main}
knitr::kable(data.frame(
  scenario = scen,
  `true Delta` = round(op$true, 1),
  `adj bias` = round(op$adj_bias, 3),
  RMSE = round(op$adj_rmse, 2),
  `95% coverage` = round(op$coverage95, 3),
  `adjusted flag (P>0.95)` = round(op$p_fav_gt_95, 3),
  `unadjusted flag (P>0.95)` = round(op$unadj_p_fav_gt_95, 3),
  check.names = FALSE),
  caption = "Operating characteristics, n = 230, CGR = 0.7, 500 trials/scenario")
```

The findings, under this generative model:

- The adjusted estimate is **approximately unbiased** (|bias| small in every
  scenario) with **near-nominal 95% coverage**.
- When the true adjusted effect is zero (rows 1 and 3), the adjusted false-favourable
  rate is **controlled** — including the pure-expectancy row, where an unadjusted
  analysis flags an effect almost every time but the adjusted analysis rarely does.
- That control has a **cost**: in the rows with a true adjusted effect (rows 2 and 4),
  the adjusted flag rate sits *below* the unadjusted flag rate in the same row —
  adjustment can **reduce detection probability**, because it discards the
  expectancy-associated part of the signal. The correction is not free.
- These results hold *under the stated generative model*; they are evidence about
  the estimator's behaviour, not proof that CGRC's causal assumptions are correct.

## Simulation as a design diagnostic

The same function is a *feasibility* check. The wrong-guess strata are populated by
roughly the $(1-\text{CGR})$ share of participants, so at a high correct-guess rate
with small $n$ a discordant treatment × guess stratum can come up empty — and the
estimand is then undefined for that trial. This is intrinsically a high-guess-rate
phenomenon: at a typical CGR of 0.75 the wrong-guess share is ~25% and empty strata
essentially never occur, whereas near CGR 0.95 the ~5% wrong-guess share is easily
exhausted. `cgr_operating()` reports how often it happens as `empty_stratum_rate`.

```{r s-sim-sparse, cache=TRUE}
sp  <- cgr_operating(n = 50, p_cg = 0.95, mu_dte = 3, mu_aeb = 7.7, n_trials = 300, seed = 1)
r10 <- sp[sp$DTE == 1 & sp$AEB == 0, ]
knitr::kable(data.frame(design = "n = 50, CGR = 0.95",
  `empty-stratum rate` = sprintf("%.0f%%", 100 * r10$empty_stratum_rate),
  `adjusted bias (feasible trials)` = round(r10$adj_bias, 2),
  `95% coverage` = round(r10$coverage95, 3), check.names = FALSE),
  caption = "Sparse-stratum diagnostic: a thin discordant stratum leaves many trials undefined")
```

A design like this — few participants, near-certain guessing — is one where CGRC
should **not** be trusted blindly: about half of trials have an empty stratum, leaving
the adjustment undefined, and even the feasible trials show degraded coverage. This is
why `cgr_operating()` is not merely a validation function for this paper; it is
something a researcher runs **before** trusting CGRC on their own trial.

## Running it on your own design

```{r s-tool-code, eval=FALSE}
# Supply a design resembling your trial and inspect the operating characteristics:
cgr_operating(n = 120, p_cg = 0.85, n_trials = 500)
```

A researcher can pass their own sample size and expected correct-guess rate and read
back bias, RMSE, interval coverage, false-favourable behaviour, and the rate at which
sparse strata make the adjustment undefined — before committing to an adjusted result.

# Extension: retaining "I do not know" guesses

Real blinding questionnaires frequently offer three options — active, placebo, or
**"I do not know"**. Dropping the UNKNOWN responses discards randomised participants,
and recoding them as active/placebo *invents* a directional belief they did not
report. **`cgrc.bayes` introduces a six-stratum extension** — not part of the
original Szigeti et al. formulation — that keeps UNKNOWN as an observed third
response category:
$$
ACAC,\ ACPL,\ ACU,\ PLAC,\ PLPL,\ PLU,
$$
(arm, then guess, with `U` = UNKNOWN). The adjusted target sets the *directional*
guess rate to chance while retaining the observed UNKNOWN-response rate,
$$
\Delta(0.50,\ u_{obs}),
$$
so the correction addresses only the informativeness of the directional guesses. At
$u=0$ this reduces exactly to the four-stratum estimand, and the observed-CGR identity
generalises: at the observed directional CGR and observed UNKNOWN rate, the reweighted
curve equals the raw arm-mean difference.

```{r s-unknown, cache=TRUE}
set.seed(3)
du  <- sim_aeb_unknown(n = 240, p_cg = 0.7, u = 0.25, dte_on = TRUE)   # a simulated 3-arm-guess trial
uf  <- cgrc_unknown(du, n_draws = 8000, direction = 1, seed = 1)
cnt <- lengths(cgr_unknown_strata(du))[UNKNOWN_STRATA]
zu  <- cgr_unknown_reference_line_test(du)
knitr::kable(data.frame(t(cnt), check.names = FALSE),
  caption = sprintf(
    "Six strata (simulated): observed directional CGR = %.3f, UNKNOWN rate = %.0f%%; identity error = %.1e",
    uf$observed_directional_cgr, 100 * uf$observed_unknown_rate,
    abs(zu$D_at_obs - zu$raw_mean_diff)))
```

This is **not** the same as assuming the UNKNOWN participants guessed incorrectly:
their uncertainty stays an observed category rather than being discarded or invented.
Section 8 applies the extension to a real trial where a third of participants answered
"I do not know."

# Worked examples on independent trial datasets

The method was not tuned to one dataset. The three trials below — spanning
psychedelics, dental sham therapy and anaesthesia-masked ketamine — are run through
**exactly the pipeline the app's *Analyse your own trial* panel uses**: audit the
inputs, detect binary vs UNKNOWN guesses, then run `cgrc()` or `cgrc_unknown()`. They
are chosen for the methodological lesson each carries, not to re-report three papers.

| Trial | Design | Guess structure | Why it is useful here |
|---|---|---|---|
| Cavanna et al. (2022) | psilocybin microdosing | binary (AC/PL) | the standard four-stratum workflow on another psychedelic dataset |
| Santana-Penín et al. (2023) | sham dental therapy | AC / PL / UNKNOWN | ~1/3 answered "I do not know" — why retaining UNKNOWN matters (six-stratum) |
| Lii et al. (2023) | ketamine masked by anaesthesia | AC / PL / UNKNOWN | guessing near chance — when blinding holds, the CGRC correction should be small |

```{r s-examples, cache=TRUE}
cav <- read_ext("cavanna_cgrc.csv")
san <- read_ext("santana_cgrc_unk.csv")
lii <- read_ext("ketamine_cgrc_unk.csv")
A <- list(
  Cavanna  = app_analyse(cav, "treatment_received", "treatment_guessed", "vas_dose_days",           direction =  1),
  Santana  = app_analyse(san, "treatment_received", "treatment_guessed", "pain_improvement_6m_itt", direction =  1, delta = 1.5),
  Lii      = app_analyse(lii, "treatment_received", "treatment_guessed", "madrs_mean_day1_3",       direction = -1))
row_of <- function(nm, a) {
  s <- a$fit$summary
  ocg <- if (a$mode == "unknown") a$fit$observed_directional_cgr else a$fit$observed_cgr
  pct <- function(p) sprintf("%.0f%%", 100 * p)
  data.frame(trial = nm, mode = a$mode, n = nrow(a$trial),
             `obs CGR` = round(ocg, 3),
             raw = round(s$post_mean[1], 2), `adjusted (0.50)` = round(s$post_mean[2], 2),
             `adj 95% CrI` = sprintf("[%.2f, %.2f]", s$cri_lo[2], s$cri_hi[2]),
             `P(fav) raw->adj` = paste(pct(s$p_favourable[1]), "->", pct(s$p_favourable[2])),
             `P(meaningful) raw->adj` = paste(pct(a$head$p_meaningful_obs), "->", pct(a$head$p_meaningful_blind)),
             check.names = FALSE)
}
knitr::kable(do.call(rbind, Map(row_of, names(A), A)), row.names = FALSE,
  caption = "Three independent trials through the app's analysis pipeline (Cavanna outcome = dosing-day VAS positive control)")
```

Read across the rows: **Santana** shows a favourable pain effect that survives
adjustment but, at the 1.5-point clinical threshold, is only borderline meaningful —
and the six-stratum analysis keeps all 77 participants rather than stranding the
third who answered UNKNOWN. **Lii** shows the opposite regime: with directional
guessing essentially at chance, the CGRC correction barely moves the estimate, which
is the correct behaviour when blinding held. **Cavanna** exercises the ordinary
four-stratum workflow on an independent psychedelic trial.

```{r s-examples-plot, fig.height=5.5, fig.show="hold", out.width="49%"}
app_curve(A$Santana, "favourable")   # UNKNOWN preserved; effect at the clinical MID
app_curve(A$Lii,     "favourable")   # near-chance guessing; correction is small
```

The same workflow behaves sensibly across qualitatively different trials. Provenance
and preparation caveats for each dataset are documented with the data and in the full
Rmd.

# From worked examples to the interactive app

The examples above are not bespoke scripts: they were run through the same analysis
logic the Shiny app exposes to users. The app makes the CGRC workflow usable by
researchers who did not develop the method and do not want to write R.

## Analyse your own trial

A researcher uploads trial-level data with columns for treatment condition, treatment
guess, and an outcome. The app then:

1. **audits** the inputs (missing values, unmappable codes, stratum counts, arm-specific
   guess rates);
2. **detects** whether guesses are binary or contain UNKNOWN responses;
3. runs the four-stratum `cgrc()` or the six-stratum `cgrc_unknown()` accordingly;
4. displays the **CGRC curve** (exactly the figures in Sections 5 and 8);
5. reports the **raw** and **adjusted** effect and the adjusted **95% credible interval**;
6. reports **$P(\text{favourable})$** and **$P(\text{meaningful})$**.

## And check whether the adjustment is trustworthy

The second panel runs the Section 6 simulation for a design the researcher describes,
returning the reliability read-out — bias, coverage, false-favourable behaviour, and
the empty-stratum rate — as a plain-language verdict. That two-part workflow is the
point:

> The analysis tells the researcher *what the CGRC adjustment is*; the simulation
> tells them *whether that adjustment is reliable for a design like theirs*.

## Try it

The whole workflow above is a thin layer over a small package API:

```{r s-app-code, eval=FALSE}
# 1. Adjust a trial (df has columns condition = AC/PL, guess = AC/PL[/UNKNOWN], value):
cgrc(my_trial)                                  # or cgrc_unknown(my_trial) when UNKNOWN is present

# 2. Check whether the adjustment is trustworthy for a design like yours:
cgr_operating(n = 120, p_cg = 0.85, n_trials = 500)

# 3. Do both interactively, no code required:
install.packages("shiny")
cgrc.bayes::cgrc_app()                          # launches the "Design" + "Analyse your own trial" app
```

`cgrc_app()` opens the interactive tool locally (see the repository README for
launch details); there is no hosted deployment. It is the most direct way to see the
whole CGRC workflow — analysis and design check together — on your own trial.

# Discussion

**What stays the same.** The Bayesian implementation does not redefine the CGRC
target: it computes exactly $\Delta(c)$ and satisfies the same observed-CGR identity.

**What the Bayesian implementation adds.** A posterior distribution over that
estimand, and direct uncertainty summaries — a credible interval, $P(\Delta>0\mid y)$
and $P(\Delta>\delta\mid y)$ — in place of resample-averaged p-values. Note the Monte
Carlo distinction precisely: the original generates its estimate by finite resampling,
whereas the Bayesian version still uses Monte Carlo draws — so it is not error-free —
but those draws *summarise a defined posterior* rather than construct the estimand;
more draws shrink summary error toward a fixed target.

**What the simulations add.** A way to examine expected bias, coverage, false-favourable
behaviour and sparse-stratum failure for a *particular* design before trusting an
adjusted result — operationalising the paper's own recommendation to simulate first.

**What the UNKNOWN extension adds.** A principled way to retain explicit "I do not know"
responses rather than discard or recode them — a measurement/data-structure fix, clearly
labelled as an addition by `cgrc.bayes`.

**What the worked examples add.** Evidence that one workflow applies across different
trial structures and behaves differently when blinding is poor, mixed, or relatively
successful.

**What the app adds.** Accessibility. A trial researcher should not have to reconstruct
the mathematics or write custom R to ask whether unblinding could plausibly alter their
result — and, just as importantly, whether the CGRC adjustment is even stable for their
design. (Retaining the posterior also enables practical-equivalence (ROPE) analyses;
those are demonstrated in the full Rmd.)

**What none of this establishes.** The central defensible claim is narrow: *the Bayesian
implementation targets the same CGRC estimand while replacing repeated pseudo-dataset
inference with a posterior over that estimand, and packages the workflow so others can
use it.* In particular:

- Attenuation is **not** a causal fraction attributable to expectancy.
- Guessing is post-randomisation and may reflect experienced efficacy or side effects,
  which the reweighting cannot disentangle.
- Reproducing the original method validates the computation, not its causal assumptions.
- The simulations validate behaviour *under the simulation model*, not real-world causal
  identification.
- The UNKNOWN extension solves a data-structure problem, not the broader causal
  limitations of CGRC.

The CGRC — in either implementation — is best read as a **sensitivity analysis for
unblinding and expectancy**, not a causal decomposition of the treatment effect.

# Sources

Only the first two references underpin this work; full per-finding citations are in
`reports/SOURCES.md`, and the external-trial datasets are cited with their archives in
`inst/extdata/external_studies/PROVENANCE.txt`.

- **Primary methodological source** — the original CGRC method reproduced here.
  Szigeti B, Nutt D, Carhart-Harris R, Erritzoe D. 2023. "The difference between
  'placebo group' and 'placebo control': a case study in psychedelic microdosing."
  *Scientific Reports* 13:12107. <https://doi.org/10.1038/s41598-023-34938-7>
- **Empirical dataset** — source of the public week-1 data reanalysed here.
  Szigeti B, et al. 2021. "Self-blinding citizen science to explore psychedelic
  microdosing." *eLife* 10:e62878. <https://doi.org/10.7554/eLife.62878>

Worked-example datasets: Cavanna et al. (2022) *Transl. Psychiatry* 12:307; Santana-Penín
et al. (2023) *Ann. Anat.* 250:152117; Lii et al. (2023) *Nat. Mental Health* 1:876–886.
This project reproduces and re-analyses published work; it is independent of and not
endorsed by the original authors.

```{r s-session, echo=FALSE}
sessionInfo()
```
