---
title: "A Bayesian implementation of the Correct Guess Rate Curve"
subtitle: "Same estimand, posterior uncertainty instead of resample-averaged p-values"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r s-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
set.seed(20260724)
options(width = 100)

# This manuscript consumes the installed cgrc.bayes package; every cgr_* and
# nig_draws() call below is the package's implementation. The complete technical
# record - full derivations, code audit, extended diagnostics, prior/robustness
# sensitivities, the UNKNOWN-response extension, the interactive app and the
# external-trial applications - lives in the exhaustive companion, CGRC_bayes.Rmd.
library(cgrc.bayes)
library(ggplot2)

# The public week-1 microdose data. Place pacutes.csv in the working directory
# (fetch + checksum-verify with data-raw/download_data.R from the repo).
stopifnot(file.exists("pacutes.csv"))
raw <- read.csv("pacutes.csv", stringsAsFactors = FALSE)

# Map the raw coding (MD = microdose/active, PL = placebo) to AC/PL and keep the
# first-week, single-timepoint rows, one record per participant.
load_scale <- function(code, tp = "w1s1") {
  d <- raw[raw$test_name == code & raw$tp == tp, ]
  data.frame(condition = ifelse(d$condition == "MD", "AC", "PL"),
             guess     = ifelse(d$guess     == "MD", "AC", "PL"),
             value     = d$value, stringsAsFactors = FALSE)
}
SCALES <- c("PANAS" = "PANAS", "Mood VAS" = "mood",
            "Energy VAS" = "energy", "Cognitive performance" = "CPS")
sb    <- lapply(SCALES, load_scale); names(sb) <- names(SCALES)
panas <- sb[["PANAS"]]

N_DRAWS  <- 20000
has_jags <- requireNamespace("rjags", quietly = TRUE) &&
            requireNamespace("coda",  quietly = TRUE)
```

# Motivation and contribution

Blinding fails in many trials, and psychedelic and microdose trials fail it
badly: participants who correctly guess their arm can respond in part to *belief*
rather than drug, and because more people in the active arm guess correctly, that
belief-driven response piles up unevenly on the treatment side. The **Correct
Guess Rate Curve (CGRC)** of Szigeti et al. (2023) asks a counterfactual of such a
trial: *what would the estimated treatment effect have been if the correct-guess
rate (CGR) had been 50% — i.e. if blinding had held?* It reweights the four
treatment × guess strata to a target guess rate while holding the within-class arm
composition fixed, so only *how much guessing happened* changes, never *who
received what*.

Szigeti et al. implement this by resampling from kernel-density estimates of the
four strata to a target guess rate, fitting a treatment contrast to each
pseudo-dataset, and averaging the resulting estimates and p-values over resamples.

This document does two things. First, it **reproduces** that estimand and its
published behaviour. Second, it presents a **Bayesian implementation of the same
estimand**: a Normal-Inverse-Gamma posterior for the four stratum means whose
uncertainty propagates analytically through the reweighting, replacing the
resample-averaged p-value with a posterior distribution over the CGR-adjusted
contrast. The estimand is unchanged; only the treatment of uncertainty differs.

The headline empirical result is that on the public microdose data the PANAS
positive-affect effect attenuates from a raw active-minus-placebo difference of
about 3.2 points to roughly 1 point once the analysis is corrected to a 50%
guess rate — consistent with a substantial expectancy contribution to the
unadjusted signal.

Two cautions frame everything below. The Bayesian version is **not** claimed to be
categorically better inference: it targets the same quantity and removes a Monte
Carlo term with no scientific content, nothing more. And **attenuation is not a
causal decomposition**: the drop from 3.2 to 1 does not measure "the fraction of
the treatment effect caused by expectancy." The CGRC is a sensitivity analysis for
unblinding under an explicit reweighting model, not a measurement of the
expectancy-mediated share of the effect.

# The Correct Guess Rate Curve estimand

Each participant contributes a treatment arm (active AC / placebo PL) and a guess
(AC / PL), giving four strata: `ACAC`, `ACPL`, `PLAC`, `PLPL` (arm, then guess).
Write $\bar y_k$ for the mean outcome in stratum $k$ and $\rho_k$ for its share of
the sample. The **observed correct-guess rate** is
$c_{obs} = \rho_{ACAC} + \rho_{PLPL}$. Two within-class ratios summarise how the
correct- and incorrect-guessers split across arms:
$$
r = \frac{\rho_{PLPL}}{\rho_{PLPL}+\rho_{ACAC}}, \qquad
s = \frac{\rho_{ACPL}}{\rho_{ACPL}+\rho_{PLAC}} .
$$
At a target guess rate $c$ the four strata receive weights
$$
w_{ACAC}=c(1-r),\quad w_{PLPL}=cr,\quad w_{ACPL}=(1-c)s,\quad w_{PLAC}=(1-c)(1-s),
$$
so the correct-guess weights sum to $c$ and the incorrect-guess weights to $1-c$,
and the CGR-adjusted contrast is the reweighted active-minus-placebo difference
$$
\Delta(c)=\frac{w_{ACAC}\bar y_{ACAC}+w_{ACPL}\bar y_{ACPL}}{w_{ACAC}+w_{ACPL}}
        -\frac{w_{PLPL}\bar y_{PLPL}+w_{PLAC}\bar y_{PLAC}}{w_{PLPL}+w_{PLAC}} .
$$
The target of interest is $\Delta(0.5)$: the effect at a 50% guess rate.

**The observed-CGR identity.** Because $r$ and $s$ are defined from the observed
composition, at $c=c_{obs}$ the reweighting is a no-op and
$$
\Delta(c_{obs}) = \bar y_{AC} - \bar y_{PL},
$$
the ordinary unadjusted arm-mean difference a trial reports in its own table. This
identity is exact and is the central validation tool used throughout: a curve is
only in the right place if it passes through the reported unadjusted effect at the
observed CGR.

```{r s-estimand}
st  <- cgr_strata(panas)                 # exact ratios (the analytic estimand path)
rat <- cgr_ratios(st)
o   <- cgr_observed(st)
w05 <- cgr_weights(0.5, rat$r, rat$s)
idn <- cgr_reference_line_test(panas, orig_cgr = o)
knitr::kable(
  data.frame(stratum = STRATA, n = lengths(st)[STRATA],
             mean = round(vapply(st[STRATA], mean, numeric(1)), 2),
             weight_at_c0.5 = round(w05[STRATA], 3)),
  caption = sprintf(
    "PANAS week 1: four strata; observed CGR = %.4f, r = %.3f, s = %.3f",
    o, rat$r, rat$s))
```

For PANAS, $\Delta(0.5) =$ `r round(cgr_delta(0.5, lapply(st, mean), rat$r, rat$s), 3)`
points, and the identity holds to machine precision:
$\Delta(c_{obs}) =$ `r round(idn$D_at_obs, 3)` equals the raw arm-mean difference
`r round(idn$raw_mean_diff, 3)` (absolute error
`r sprintf("%.1e", abs(idn$D_at_obs - idn$raw_mean_diff))`). That reweighted
1-point value at perfect blinding, against the ~3.2-point raw difference, is the
CGRC's substantive claim for PANAS.

The reproduction of Szigeti's *shipped* code uses a rounded-ratio variant
(`legacy_round = TRUE`), which shifts these numbers by hundredths of a point; the
estimand itself uses the exact ratios above. The two paths are kept distinct and
are compared in the full Rmd.

# Bayesian implementation of the same estimand

The original procedure obtains uncertainty by resampling pseudo-datasets. The
Bayesian implementation instead places a posterior on each stratum mean and
propagates it through $\Delta(c)$ directly. Each stratum is modelled as
$y_{ij}\mid\mu_j,\sigma^2_j \sim N(\mu_j,\sigma^2_j)$ with a
Normal-Inverse-Gamma prior
$$
\mu_j\mid\sigma^2_j \sim N(m_0,\ \sigma^2_j/k_0),\qquad
\sigma^2_j \sim \text{Inv-Gamma}(a_0,b_0),
$$
whose four hyperparameters have plain meanings: $m_0$ is the prior mean, $k_0$ a
prior sample size in pseudo-observations, and $(a_0,b_0)$ the prior shape and
scale for the variance. With the vague defaults used here ($m_0=0$, $k_0=10^{-6}$,
$a_0=b_0=10^{-3}$) the prior contributes negligibly against dozens of real
observations. The conjugate update is closed-form,
$$
k_n=k_0+n,\quad m_n=\frac{k_0 m_0+n\bar y}{k_n},\quad a_n=a_0+\tfrac n2,\quad
b_n=b_0+\tfrac12\!\sum_i (y_i-\bar y)^2+\frac{k_0 n(\bar y-m_0)^2}{2k_n},
$$
giving $\sigma^2_j\mid y \sim \text{Inv-Gamma}(a_n,b_n)$ and
$\mu_j\mid\sigma^2_j,y \sim N(m_n,\sigma^2_j/k_n)$.

Posterior draws of the four stratum means are pushed through $\Delta(c)$ for every
$c$, yielding a posterior over the whole curve — a posterior mean, a 95% credible
interval, and the posterior probability that the effect is favourable, at each
guess rate. The number of draws controls only Monte Carlo error in these
posterior summaries; it is not a sample size and carries no information about
participants. (The order of the conjugate draws, the within-draw dependence of
$\mu$ and $\sigma^2$, and the full derivation are in the complete Rmd; they are not
needed to read the results.)

```{r s-bayes}
fit_panas <- cgrc(panas, n_draws = N_DRAWS, seed = 2)   # posterior propagation
knitr::kable(
  fit_panas$summary[, c("what", "post_mean", "cri_lo", "cri_hi", "p_favourable")],
  digits = 3,
  caption = "PANAS: CGR-adjusted posterior (Normal-Inverse-Gamma conjugate)")
```

# Reproduction and validation

## Reproduction of the published CGRC

Under the activated-expectancy-bias (AEB) generative model of Szigeti et al., the
implementation reproduces the published simulation behaviour. At the paper's regime
($n=230$, CGR $=0.7$, a 3-point effect, expectancy magnitude 7.7), the unadjusted
two-sided significance rates across the four (effect, expectancy) scenarios match
the published Table 1 values (0.05 / 0.86 / 0.78 / 0.99), and the CGR-adjusted
estimator is close to unbiased with near-nominal coverage.

```{r s-repro, cache=TRUE}
op <- cgr_operating(n = 230, p_cg = 0.7, mu_dte = 3, mu_aeb = 7.7,
                    n_trials = 500, seed = 1)
knitr::kable(
  data.frame(
    scenario = c("no effect / no expectancy", "effect / no expectancy",
                 "no effect / expectancy", "effect / expectancy"),
    `unadjusted p<0.05 (2-sided)` = round(op$freq_sig, 3),
    `adjusted bias` = round(op$adj_bias, 3),
    `95% coverage` = round(op$coverage95, 3),
    check.names = FALSE),
  caption = "AEB simulation, n = 230, CGR = 0.7, 500 trials (published Table 1: 0.05/0.86/0.78/0.99)")
```

## The observed-CGR identity

The identity of Section 2 holds for every outcome: the CGR-adjusted curve recovers
the ordinary unadjusted contrast at the observed CGR, to machine precision. This is
what makes a reference line drawn on the curve checkable rather than decorative.

```{r s-identity}
knitr::kable(
  do.call(rbind, lapply(names(sb), function(nm) {
    z <- cgr_reference_line_test(sb[[nm]], orig_cgr = cgr_observed(cgr_strata(sb[[nm]])))
    data.frame(outcome = nm, observed_CGR = round(z$computed_obs_cgr, 4),
               `Delta(c_obs)` = round(z$D_at_obs, 3),
               `raw AC-PL`    = round(z$raw_mean_diff, 3),
               `abs error`    = sprintf("%.1e", abs(z$D_at_obs - z$raw_mean_diff)),
               check.names = FALSE)
  })),
  caption = "Observed-CGR identity across the four outcomes")
```

## KDE resampling versus the analytic estimand

The original kernel-density resampling and the analytic $\Delta(c)$ target the same
quantity. As the resample count grows, the KDE average converges on the analytic
value; the only systematic difference at small counts is Monte Carlo variation.
Because $\Delta(c)$ depends on the strata only through their means, and
kernel-density smoothing preserves means, the KDE step does not materially change
the estimand.

```{r s-kde, cache=TRUE}
set.seed(11)
kl <- cgr_kde_ladder(panas, cgr_observed(cgr_strata(panas)),
                     reps = c(32, 100, 1000, 10000))
knitr::kable(kl[, c("n_rep", "kde_est", "kde_mcse", "analytic", "diff_from_analytic")],
             digits = 3,
             caption = "KDE resampling converges on the analytic Delta (PANAS, observed CGR)")
```

## Original inference versus Bayesian inference

The two implementations agree on the estimand but differ in their summary of
evidence. Over repeated simulated trials, the KDE point estimate and the Bayesian
posterior mean of $\Delta(0.5)$ are almost perfectly correlated and share the same
small bias. Their *decisions*, however, differ: at matched thresholds the posterior
flags the effect appreciably more often than a rule based on the resample-averaged
p-value, and the two disagree on a fifth of trials.

```{r s-headtohead, cache=TRUE}
set.seed(101)
N <- 300; n_i <- 200; p_cg <- 0.7; mu_dte <- 3
ke <- kp <- be <- bp <- rep(NA_real_, N)
for (i in seq_len(N)) {
  di <- sim_aeb(n_i, p_cg, dte_on = TRUE, aeb_on = FALSE, mu_dte = mu_dte)
  if (!all(STRATA %in% paste0(di$condition, di$guess))) next
  ki <- cgr_kde(di, 0.5, n_rep = 100); ke[i] <- ki$est; kp[i] <- ki$p
  sti <- cgr_strata(di); rti <- cgr_ratios(sti)          # loop-local; do not clobber PANAS st/rat
  ddi <- cgr_delta(0.5, lapply(sti, nig_draws, n_draws = 4000), rti$r, rti$s)
  be[i] <- mean(ddi); bp[i] <- mean(ddi > 0)
}
ok <- !is.na(kp)
knitr::kable(data.frame(
  Quantity = c("Point-estimate correlation (KDE vs posterior)",
               "Point-estimate bias — KDE / posterior",
               "Flag rate — resample-averaged p < 0.05",
               "Flag rate — posterior P(favourable) > 0.975",
               "Trials where the two decisions agree"),
  Value = c(sprintf("%.3f", cor(ke[ok], be[ok])),
            sprintf("%+.3f / %+.3f", mean(ke[ok]) - mu_dte, mean(be[ok]) - mu_dte),
            sprintf("%.3f", mean(kp[ok] < 0.05)),
            sprintf("%.3f", mean(bp[ok] > 0.975)),
            sprintf("%.0f%%", 100 * mean((kp[ok] < 0.05) == (bp[ok] > 0.975)))),
  check.names = FALSE),
  caption = "KDE procedure vs Bayesian posterior, 300 simulated trials (true Delta = 3)")
```

Averaging a p-value across resampled pseudo-datasets does not produce a conventional
p-value: the average has no accepted sampling interpretation, and here it discards
detection power that the posterior retains. (In the full Rmd a bandwidth-zero control
confirms this is the averaging step, not the kernel.) Retaining the posterior draws
instead yields directly interpretable quantities — the point of Section 5.

# Empirical microdosing application

The public data are the self-blinding microdose trial re-analysed by Szigeti et al.
(2023). Following the paper, only week-1, single-timepoint records are used, giving
one observation per participant. For PANAS this is `r nrow(panas)` participants with
an observed CGR of `r round(o, 3)`.

The public week-1 data give an overall CGR of 0.647, whereas Figure 4 marks 0.72.
Inspection of the original source code shows that 0.72 is a hardcoded plotting
constant rather than the CGR calculated from the analyzed data. The CGRC identity
confirms that 0.647, not 0.72, is the point at which the curve recovers the reported
unadjusted effect.

```{r s-empirical}
emp <- do.call(rbind, lapply(names(sb), function(nm) {
  d <- sb[[nm]]
  f <- cgrc(d, n_draws = N_DRAWS, seed = 2)
  h <- cgrc_headline(d, n_draws = N_DRAWS, seed = 2)   # delta = 0.5 SD MID
  s <- f$summary
  data.frame(outcome = nm, n = nrow(d), observed_CGR = round(f$observed_cgr, 3),
             `raw effect` = round(s$post_mean[1], 2),
             `adjusted (CGR 0.50)` = round(s$post_mean[2], 2),
             `95% CrI` = sprintf("[%.2f, %.2f]", s$cri_lo[2], s$cri_hi[2]),
             `P(favourable)` = sprintf("%.0f%%", 100 * s$p_favourable[2]),
             `P(meaningful)` = sprintf("%.0f%%", 100 * h$p_meaningful_blind),
             check.names = FALSE)
}))
knitr::kable(emp, caption = "CGR-adjusted results for the four outcomes (adjusted = perfect blinding, CGR 0.50; meaningful = beyond a half-SD difference)")
```

```{r s-panas-fig, fig.height=6}
plot(fit_panas, title = "PANAS, week 1")
```

The table reproduces the published attenuation pattern: PANAS and the mood/energy
VAS effects shrink substantially once corrected to a 50% guess rate, with the
cognitive-performance outcome near zero throughout.

Two probabilities carry the practical message, and they are **not** the same
quantity. $P(\text{favourable})$ — the posterior probability the adjusted effect is
on the beneficial side of zero — answers "is there an effect at all." $P(\text{
meaningful})$ — the posterior probability the effect exceeds a pre-declared
minimally-important difference (a half standard deviation here) — answers "is it
large enough to matter." A tail probability is not a p-value: it is a direct
statement about the parameter given the data, and separating "non-zero" from
"clinically important" is exactly what a single significance verdict cannot do.

# Simulation and robustness

The AEB operating-characteristics study (Section 4.1, `cgr_operating`) establishes
the estimator's behaviour under the stated model: the CGR-adjusted estimate is
approximately unbiased with near-nominal 95% credible-interval coverage, and when
the true adjusted effect is zero the false-favourable rate is controlled. Two
caveats matter for design. CGR adjustment can **reduce power** relative to the
unadjusted comparison, because it discards the expectancy-driven part of the signal.
And performance **deteriorates when strata become sparse**: at a high guess rate with
small $n$, a wrong-guess stratum can be empty and the estimand undefined for that
trial.

```{r s-sim-sparse, cache=TRUE}
sp <- cgr_operating(n = 80, p_cg = 0.9, mu_dte = 3, mu_aeb = 7.7,
                    n_trials = 300, seed = 1)
r11 <- sp[sp$DTE == 1 & sp$AEB == 0, ]
knitr::kable(
  data.frame(design = "n = 80, CGR = 0.90",
             `empty-stratum rate` = sprintf("%.0f%%", 100 * r11$empty_stratum_rate),
             `adjusted bias (feasible trials)` = round(r11$adj_bias, 2),
             `95% coverage` = round(r11$coverage95, 3),
             check.names = FALSE),
  caption = "Sparse-stratum deterioration: a thin discordant stratum leaves many trials undefined")
```

**Robustness to the within-stratum model.** Refitting with a Student-t likelihood
(a heavier-tailed alternative to the normal) probes sensitivity to the Gaussian
assumption. The exact attenuation magnitude shifts a little, but the qualitative
inference does not: the estimated degrees of freedom are large, so the t collapses
toward the normal. This is why *percentage* attenuation should be read as
model-dependent while the direction and rough size of the correction are stable.

```{r s-robust, eval=has_jags, cache=TRUE}
set.seed(6)
jn <- cgr_jags(panas, likelihood = "normal", n_iter = 10000, n_chains = 4)
jt <- cgr_jags(panas, likelihood = "t",      n_iter = 10000, n_chains = 4)
i5 <- function(x) which.min(abs(x$cgr - 0.5))
knitr::kable(data.frame(
  likelihood = c("normal", "Student-t"),
  adjusted_at_0.5 = round(c(jn$est[i5(jn)], jt$est[i5(jt)]), 3),
  cri_95 = c(sprintf("[%.2f, %.2f]", jn$lo[i5(jn)], jn$hi[i5(jn)]),
             sprintf("[%.2f, %.2f]", jt$lo[i5(jt)], jt$hi[i5(jt)])),
  P_favourable = round(c(jn$p_fav[i5(jn)], jt$p_fav[i5(jt)]), 3),
  est_df_nu = c(NA, round(attr(jt, "nu"), 1))),
  caption = "PANAS at CGR 0.50: normal vs Student-t likelihood (JAGS)")
```

```{r s-robust-note, echo=FALSE, results='asis'}
if (!has_jags) cat(
  "*The Student-t robustness check requires a JAGS install (`rjags`), which is not",
  "available in this render. In the complete Rmd it runs and gives an estimated",
  "nu of roughly 18, so the t collapses toward the normal and the Gaussian",
  "conclusion is robust.*\n")
```

Additional sensitivity analyses — a corrected JAGS backend cross-check, prior
sensitivity, Beta uncertainty in $r$ and $s$, per-stratum skewness/kurtosis, and
small-stratum subsampling — were conducted and are retained in the complete Rmd.
Their qualitative message is the same: the estimand and the direction of the
correction are stable; only the exact magnitude is model-dependent.

# Discussion

**Same estimand.** The Bayesian implementation does not redefine the CGRC target.
It computes exactly $\Delta(c)$ — the reweighted active-minus-placebo contrast at a
target guess rate — and satisfies the same observed-CGR identity.

**A different treatment of uncertainty.** Where the original generates uncertainty
by repeatedly resampling pseudo-datasets, the Bayesian version places a posterior on
the four stratum means and propagates it analytically through the reweighting. On
the Monte Carlo distinction, precision matters: the original introduces Monte Carlo
error by *generating the estimate* from finite resampling, whereas the Bayesian
implementation still uses Monte Carlo draws — so it is not error-free — but those
draws merely *summarise a defined posterior distribution* rather than construct the
estimand. More draws shrink summary error toward a fixed target; more resamples
chase a moving one.

**Interpretability.** Retaining the posterior gives directly interpretable
quantities such as $P(\Delta>0\mid y)$ and $P(\Delta>\delta\mid y)$ — probability of
a favourable effect and probability of a meaningful one — without averaging p-values
across pseudo-datasets. Retaining the draws also enables practical-equivalence
(ROPE) decompositions of the posterior around a clinical threshold; that machinery
and its width-sensitivity analysis are demonstrated in the complete Rmd.

**What this does not establish.** The central defensible claim is narrow: *the
Bayesian implementation targets the same CGRC estimand while replacing repeated
pseudo-dataset inference with a posterior distribution over that estimand.* In
particular:

- Attenuation is **not** the percentage of the treatment effect caused by
  expectancy; the drop from ~3.2 to ~1 on PANAS is not a causal share.
- The CGRC depends on the assumptions of its reweighting estimand (fixed
  within-class arm ratios, balanced allocation, the AEB generative model in the
  simulations); reproducing the original curve validates the computation, not those
  assumptions.
- Guessing may itself be influenced by experienced efficacy, side effects or other
  post-randomisation information, which the reweighting cannot disentangle.
- The analysis is best read as a **sensitivity analysis for unblinding and
  expectancy**, not a causal decomposition of the treatment effect.

# Sources

Only the first two references underpin this work; full per-finding citations are in
`reports/SOURCES.md`.

- **Primary methodological source** — the original CGRC method reproduced here.
  Szigeti B, Nutt D, Carhart-Harris R, Erritzoe D. 2023. "The difference between
  'placebo group' and 'placebo control': a case study in psychedelic microdosing."
  *Scientific Reports* 13:12107. <https://doi.org/10.1038/s41598-023-34938-7>
- **Empirical dataset** — source of the public week-1 data reanalysed here.
  Szigeti B, et al. 2021. "Self-blinding citizen science to explore psychedelic
  microdosing." *eLife* 10:e62878. <https://doi.org/10.7554/eLife.62878>

This project reproduces and re-analyses published work; it is independent of and not
endorsed by the original authors.

```{r s-session, echo=FALSE}
sessionInfo()
```
