---
title: "The Correct Guess Rate Curve as a usable analysis tool"
subtitle: "Rebuilding a blinding-adjustment method on firmer statistical footing, extending it to real-world 'I don't know' answers, stress-testing it by simulation, and wrapping it in a point-and-click app"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: kable
---

```{r s-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
set.seed(20260724)
options(width = 100)

# This manuscript consumes the installed cgrc.bayes package; every cgr_* /
# cgrc*/ nig_draws() call below is the package's implementation. The exhaustive
# technical record - full derivations, code audit, extended diagnostics, prior
# and robustness sensitivities, ROPE analyses and the complete external-data
# provenance - is the companion document CGRC_bayes.Rmd. This summary is the
# focused story.
library(cgrc.bayes)
library(ggplot2)

# --- primary data: the public week-1 microdose trial ------------------------
# Put pacutes.csv in the working directory (fetch + checksum-verify with
# data-raw/download_data.R). MD = microdose/active, PL = placebo.
stopifnot(file.exists("pacutes.csv"))
raw <- read.csv("pacutes.csv", stringsAsFactors = FALSE)
load_scale <- function(code, tp = "w1s1") {
  d <- raw[raw$test_name == code & raw$tp == tp, ]
  data.frame(condition = ifelse(d$condition == "MD", "AC", "PL"),
             guess     = ifelse(d$guess     == "MD", "AC", "PL"),
             value     = d$value, stringsAsFactors = FALSE)
}
SCALES <- c("PANAS" = "PANAS", "Mood VAS" = "mood",
            "Energy VAS" = "energy", "Cognitive performance" = "CPS")
sb    <- lapply(SCALES, load_scale); names(sb) <- names(SCALES)
panas <- sb[["PANAS"]]

# --- external trial datasets ------------------------------------------------
# Put these CSVs in the working directory alongside pacutes.csv.
read_ext <- function(f) {
  stopifnot(file.exists(f))
  utils::read.csv(f, stringsAsFactors = FALSE)
}

# --- the app's "Analyse your own trial" logic, as a helper ------------------
# Audit inputs, detect binary vs UNKNOWN guesses, then run the four-stratum
# cgrc() or the six-stratum cgrc_unknown(). This is the same decision the Shiny
# app makes; reused here so the worked examples are literally the app workflow.
app_analyse <- function(df, condc, guessc, valc, direction = 1, delta = NULL,
                        delta_sd_frac = 0.5, n_draws = 8000) {
  aud   <- cgrc_input_audit(df[[condc]], df[[guessc]], df[[valc]])
  clean <- aud$clean
  mode  <- if (aud$has_unknown) "unknown" else "binary"
  if (mode == "binary") clean <- clean[clean$guess %in% c("AC", "PL"), , drop = FALSE]
  d <- if (!is.null(delta)) delta else delta_sd_frac * stats::sd(clean$value)
  fit <- if (mode == "unknown")
           cgrc_unknown(clean, n_draws = n_draws, direction = direction, seed = 1)
         else cgrc(clean, n_draws = n_draws, direction = direction, seed = 1)
  hd  <- if (mode == "unknown")
           cgrc_unknown_headline(clean, direction = direction, delta = d, n_draws = n_draws, seed = 1)
         else cgrc_headline(clean, direction = direction, delta = d, n_draws = n_draws, seed = 1)
  list(mode = mode, audit = aud, trial = clean, dir = direction, delta = d, fit = fit, head = hd)
}
app_curve <- function(a, label) {
  if (a$mode == "unknown")
    cgr_unknown_plot(a$fit$curve, obs_cgr = a$fit$observed_directional_cgr,
                     u = a$fit$target_unknown_rate, direction_label = label)
  else cgr_plot(a$fit$curve, obs_cgr = a$fit$observed_cgr, direction_label = label)
}

N_DRAWS  <- 20000
has_jags <- requireNamespace("rjags", quietly = TRUE) &&
            requireNamespace("coda",  quietly = TRUE)
```

*How to read this document.* It is written to be followed without a statistics
background: every technical term is explained the first time it appears, and the
plain-English story is complete on its own. All the R code that produces the tables
and figures is hidden behind the grey **Code** buttons on the right — you can click
any of them to see exactly how a number was computed, but you never need to. Nothing
below is typed in by hand; every figure is generated from the data and the package.

# The problem, in plain terms

Clinical trials rely on *blinding*: neither the participant nor, ideally, anyone
assessing them should know who received the real treatment and who received a
placebo (a dummy with no active ingredient). Blinding matters because people who
*believe* they got the real drug often feel better simply from that belief — the
well-known placebo effect. If participants can tell which group they are in, that
belief-driven response leaks into the results and inflates the apparent benefit of
the drug.

In psychedelic and microdosing trials, blinding fails badly. The drug's effects are
often noticeable, so participants frequently guess correctly which arm they were in.
Crucially, the people who got the *real* drug guess correctly more often than the
people who got placebo. That imbalance means the belief-driven boost piles up
unevenly, mostly on the treatment side, and masquerades as a genuine drug effect.

Szigeti and colleagues (2023) proposed a clever way to gauge how much this matters.
Their **Correct Guess Rate Curve (CGRC)** asks a "what if" question:

> *If blinding had actually worked — if participants had been no better than a coin
> flip at guessing their arm — what would the estimated treatment effect have looked
> like?*

The "correct guess rate" (CGR) is just the fraction of participants who guessed
their arm correctly. Perfect blinding corresponds to a CGR of 50% (pure guessing).
The CGRC recomputes the drug-versus-placebo difference *as if* the guess rate had
been 50%, and compares that adjusted number to the raw, unadjusted one. If the effect
shrinks a lot when you "turn off" the correct guessing, that is a warning sign that
unblinding may be propping up the result.

**This package, `cgrc.bayes`, does four things with that idea**, all aimed at making
it something a working researcher can actually use on a real trial:

> It faithfully **reproduces** the original method; it **rebuilds** the same
> calculation on a cleaner statistical footing that gives directly interpretable
> uncertainty; it **stress-tests** the method by simulation so you can tell how well
> it performs *for your particular trial design*; it **extends** the method to
> the common real-world case where participants are allowed to answer "I don't know";
> and it **packages** the whole workflow into a point-and-click app, demonstrated on
> several independent trials.

**One caution frames everything below.** The amount by which an effect shrinks is
**not** a measure of "how much of the drug's effect was really just belief." The
CGRC is a *sensitivity analysis* — a structured way of asking "how fragile is this
result to unblinding?" — not a clean causal accounting. We return to this repeatedly,
because it is the easiest thing to over-claim.

# The core idea: "what if blinding had held?"

Here is the machinery, built up in plain steps.

Every participant gives us two facts: which arm they were actually in — **active
(AC)** or **placebo (PL)** — and which arm they *guessed* they were in (again AC or
PL). Crossing those two facts gives four groups, which we will call **strata**
(statistics jargon for "subgroups"). Writing arm first and guess second, they are:

- `ACAC` — got the drug, guessed drug (a correct guess),
- `ACPL` — got the drug, guessed placebo (wrong),
- `PLAC` — got placebo, guessed drug (wrong),
- `PLPL` — got placebo, guessed placebo (correct).

For each stratum we record its average outcome (say, its average mood score),
written $\bar y_k$, and what fraction of all participants fall into it, written
$\rho_k$. The **observed correct-guess rate** is simply the share of people who
guessed right: $c_{obs} = \rho_{ACAC} + \rho_{PLPL}$.

Two more quantities describe *how the correct-guessers and the wrong-guessers split
between the two arms* — because that split is exactly what the belief effect
distorts:
$$
r=\frac{\rho_{PLPL}}{\rho_{PLPL}+\rho_{ACAC}},\qquad
s=\frac{\rho_{ACPL}}{\rho_{ACPL}+\rho_{PLAC}}.
$$
In words: $r$ is the fraction of the *correct*-guessers who were in the placebo arm,
and $s$ is the fraction of the *wrong*-guessers who got the drug.

Now the "what if." To imagine a world with a chosen guess rate $c$, we simply
**re-weight** the four strata — give them new relative importances that add up to a
guess rate of $c$ instead of $c_{obs}$ — while keeping each side's internal balance
fixed. The weights are $w_{ACAC}=c(1-r)$, $w_{PLPL}=cr$, $w_{ACPL}=(1-c)s$, and
$w_{PLAC}=(1-c)(1-s)$. With those weights the **adjusted treatment effect** is the
re-weighted "drug average minus placebo average":
$$
\Delta(c)=\frac{w_{ACAC}\bar y_{ACAC}+w_{ACPL}\bar y_{ACPL}}{w_{ACAC}+w_{ACPL}}
        -\frac{w_{PLPL}\bar y_{PLPL}+w_{PLAC}\bar y_{PLAC}}{w_{PLPL}+w_{PLAC}}.
$$
Read $\Delta(c)$ as "the estimated drug effect if the correct-guess rate had been
$c$." The number we care about is $\Delta(0.5)$ — the effect under perfect blinding
(a 50% guess rate).

There is one elegant self-check hidden in this. Because $r$ and $s$ are read off the
*actual* data, if you set $c$ equal to the *actual* guess rate $c_{obs}$, the
re-weighting changes nothing, and the formula collapses back to the ordinary,
unadjusted difference in arm averages:
$$
\Delta(c_{obs})=\bar y_{AC}-\bar y_{PL}.
$$
We call this the **observed-CGR identity**, and it is the cleanest correctness test
in the whole project: any correctly implemented curve *must* pass exactly through the
plain unadjusted effect at the observed guess rate. If it doesn't, there is a bug.

The table below shows the four strata for the real PANAS mood outcome (from the
public microdosing dataset), along with the observed guess rate and the $r$ and $s$
splits:

```{r s-estimand}
st  <- cgr_strata(panas); rat <- cgr_ratios(st); o <- cgr_observed(st)
idn <- cgr_reference_line_test(panas, orig_cgr = o)
knitr::kable(
  data.frame(stratum = STRATA, n = lengths(st)[STRATA],
             mean = round(vapply(st[STRATA], mean, numeric(1)), 2)),
  caption = sprintf("PANAS week 1: four strata; observed CGR = %.4f, r = %.3f, s = %.3f",
                    o, rat$r, rat$s))
```

The identity checks out to the limit of the computer's arithmetic: the adjusted
effect at the observed guess rate, $\Delta(c_{obs})=$
`r round(idn$D_at_obs, 3)`, equals the raw difference in arm averages,
`r round(idn$raw_mean_diff, 3)`, with a discrepancy of only
`r sprintf("%.1e", abs(idn$D_at_obs - idn$raw_mean_diff))` (essentially zero). And
the adjusted effect under *perfect* blinding is
$\Delta(0.5)=$ `r round(cgr_delta(0.5, lapply(st, mean), rat$r, rat$s), 3)`.

That is the CGRC's substantive claim for PANAS in one comparison: an effect of about
**3.2 points** as measured, but only about **1 point** once you imagine away the
correct guessing. (A small technical footnote: reproducing Szigeti's *shipped* code
requires a rounded-ratio variant, kept in the package as `legacy_round = TRUE`; the
estimand itself uses the exact ratios above. The two are kept separate and compared
in the full technical document, `CGRC_bayes.Rmd`.)

# Our version: keeping the full range of plausible answers

The original method produces its estimate — and its uncertainty — by a computational
trick called *resampling*: it repeatedly manufactures artificial "pseudo-datasets"
that resemble the real strata, recomputes the effect on each, and averages the
results. Averaging the *estimates* this way is fine. But the original also averages
**p-values** across those pseudo-datasets, and a p-value has no accepted meaning once
you average it over invented data. This is the piece we wanted to put on firmer
ground.

**A one-paragraph primer on the two ways to express uncertainty.** A *p-value*
answers a deliberately roundabout question: "if the drug truly did nothing, how
surprising would data like ours be?" It is easy to misread and, once averaged over
fabricated datasets, not well defined. The Bayesian alternative we use answers the
*direct* question instead. Rather than committing to a single best-guess number, it
keeps the **whole range of effect sizes that are consistent with the data**, each
weighted by how plausible it is. That weighted range is called the **posterior
distribution**. From it you can read, in plain language:

- a single best estimate (the posterior mean),
- a **credible interval** — a range that genuinely has, say, a 95% chance of
  containing the true effect (this is the intuitive meaning people *wrongly* attach
  to the classical "confidence interval"; here it is actually correct),
- and the **probability the drug helped** — literally the share of the plausible
  range that lies on the beneficial side of zero.

Mechanically, we put a standard, deliberately weak statistical model on each
stratum's average (a "Normal-Inverse-Gamma" model — the workhorse conjugate model for
a mean and variance). "Weak" here means the model barely nudges the answer: with
dozens of real observations per stratum, the data do essentially all the talking. The
key move is then simple:

> Draw many plausible values for the four stratum averages from the posterior; for
> each draw, run them through the *same* CGRC formula $\Delta(c)$ from the previous
> section. The result is not one curve but a whole cloud of plausible curves — from
> which we read a best-estimate curve, a 95% credible band around it, and, at every
> guess rate, the probability the effect is favourable.

Two practical notes. First, the "number of draws" is just how finely we sample that
cloud; more draws sharpen the picture but are **not** extra data and cannot
manufacture certainty. Second, everything downstream is the ordinary CGRC estimand —
we changed *how uncertainty is expressed*, not *what is being estimated*.

Here is that posterior summary for PANAS, at the observed guess rate ("raw") and
under perfect blinding ("adjusted"):

```{r s-bayes}
fit_panas <- cgrc(panas, n_draws = N_DRAWS, seed = 2)
knitr::kable(fit_panas$summary[, c("what", "post_mean", "cri_lo", "cri_hi", "p_favourable")],
             digits = 3, caption = "PANAS: CGR-adjusted posterior (Normal-Inverse-Gamma conjugate)")
```

Two probabilities carry the practical message, and they answer **different**
questions:

- **$P(\text{favourable})$** — the probability the adjusted effect is on the
  beneficial side of zero — answers *"is there an effect at all?"*
- **$P(\text{meaningful})$** — the probability the effect is bigger than a
  pre-declared "smallest difference worth caring about" — answers *"is it large
  enough to matter clinically?"*

Both are direct statements about the effect given the data, which is exactly what a
clinician or reviewer actually wants to know.

# Checking it against the original method

Before trusting a rebuild, you have to show it reproduces the original where the two
should agree. Four checks do that.

**Check 1 — We reproduce the published behaviour.** Szigeti et al. characterised
their method by simulating trials under a specific model of expectancy bias (they
call it "activated expectancy bias," AEB — the belief-driven boost switched on for
correct-guessers). Re-running that simulation at the paper's settings (230
participants, a 70% guess rate, a genuine 3-point drug effect) reproduces their
published rates of "statistically significant" results across four scenarios (no
effect / a real effect / an expectancy confound / both). The numbers match the
paper's Table 1 (0.05, 0.86, 0.78, 0.99). The very same 500-trial run is reused as
the main stress-test in a later section, so nothing here is cherry-picked.

```{r s-repro, cache=TRUE}
op <- cgr_operating(n = 230, p_cg = 0.7, mu_dte = 3, mu_aeb = 7.7,
                    n_trials = 500, seed = 1)
scen <- c("no effect / no expectancy", "effect / no expectancy",
          "no effect / expectancy", "effect / expectancy")
knitr::kable(data.frame(scenario = scen,
  `unadjusted p<0.05 (2-sided)` = round(op$freq_sig, 3),
  `published Table 1` = c(0.05, 0.86, 0.78, 0.99), check.names = FALSE),
  caption = "AEB simulation, n = 230, CGR = 0.7, 500 trials: reproduces the published significance rates")
```

**Check 2 — The self-test passes for every outcome.** Recall the observed-CGR
identity: at the actual guess rate, the adjusted curve must land exactly on the plain
unadjusted effect. It does, for all four outcomes, down to the last decimal the
computer can carry (the "abs error" column is essentially zero):

```{r s-identity}
knitr::kable(do.call(rbind, lapply(names(sb), function(nm) {
  z <- cgr_reference_line_test(sb[[nm]], orig_cgr = cgr_observed(cgr_strata(sb[[nm]])))
  data.frame(outcome = nm, observed_CGR = round(z$computed_obs_cgr, 4),
             `Delta(c_obs)` = round(z$D_at_obs, 3), `raw AC-PL` = round(z$raw_mean_diff, 3),
             `abs error` = sprintf("%.1e", abs(z$D_at_obs - z$raw_mean_diff)),
             check.names = FALSE)
})), caption = "Observed-CGR identity across the four outcomes")
```

**Check 3 — The original's resampling homes in on the exact answer.** If you let the
original method's resampling run longer and longer (more and more pseudo-datasets),
its averaged estimate converges on the exact value our formula computes directly. The
gap at small resample counts is just random noise from the resampling, not a real
disagreement. (The reason is that the CGRC only ever uses the strata's *averages*,
and the resampling smoothing leaves those averages untouched.)

```{r s-kde, cache=TRUE}
set.seed(11)
kl <- cgr_kde_ladder(panas, cgr_observed(cgr_strata(panas)), reps = c(32, 100, 1000, 10000))
knitr::kable(kl[, c("n_rep", "kde_est", "kde_mcse", "analytic", "diff_from_analytic")],
             digits = 3, caption = "KDE resampling converges on the analytic Delta (PANAS, observed CGR)")
```

**Check 4 — The two methods agree on the estimate, and differ only in how they
express uncertainty.** Running both methods on the same 300 simulated trials, their
*point estimates* are almost perfectly correlated (correlation ≈ 1) with the same
tiny bias. Where they part ways is the yes/no *decision*, precisely because the
original leans on an averaged p-value that has no clean interpretation.

```{r s-headtohead, cache=TRUE}
set.seed(101); N <- 300; n_i <- 200; p_cg <- 0.7; mu_dte <- 3
ke <- kp <- be <- bp <- rep(NA_real_, N)
for (i in seq_len(N)) {
  di <- sim_aeb(n_i, p_cg, dte_on = TRUE, aeb_on = FALSE, mu_dte = mu_dte)
  if (!all(STRATA %in% paste0(di$condition, di$guess))) next
  ki <- cgr_kde(di, 0.5, n_rep = 100); ke[i] <- ki$est; kp[i] <- ki$p
  sti <- cgr_strata(di); rti <- cgr_ratios(sti)
  ddi <- cgr_delta(0.5, lapply(sti, nig_draws, n_draws = 4000), rti$r, rti$s)
  be[i] <- mean(ddi); bp[i] <- mean(ddi > 0)
}
ok <- !is.na(kp)
knitr::kable(data.frame(
  Quantity = c("Point-estimate correlation (KDE vs posterior)",
               "Point-estimate bias: KDE / posterior",
               "Flag rate: KDE resample-averaged two-sided p < 0.05",
               "Flag rate: posterior P(favourable) > 0.975 (matches two-sided 0.05)",
               "Trials where the two decisions agree"),
  Value = c(sprintf("%.3f", cor(ke[ok], be[ok])),
            sprintf("%+.3f / %+.3f", mean(ke[ok]) - mu_dte, mean(be[ok]) - mu_dte),
            sprintf("%.3f", mean(kp[ok] < 0.05)),
            sprintf("%.3f", mean(bp[ok] > 0.975)),
            sprintf("%.0f%%", 100 * mean((kp[ok] < 0.05) == (bp[ok] > 0.975)))),
  check.names = FALSE),
  caption = "Original vs Bayesian inference, 300 simulated trials (true Delta = 3)")
```

The takeaway: the two approaches agree closely on the point estimand but can produce
materially different evidence summaries. Under the matched thresholds used here, the
posterior rule flags the effect more often than the resample-averaged p-value rule.
This is a difference in how the evidence is *summarised* — the posterior states a
probability about the effect directly, in place of an averaged p-value — not a claim
that one method is universally better.

*A note on the "0.975" in that table, since it looks like an odd number.* It is
there only to make the comparison fair. The original method's decision rule is a
**two-sided** significance test at the 5% level. A two-sided 5% test splits its 5%
into 2.5% on each side, so the matching one-sided probability bar is
$1 - 0.025 = 0.975$. Using that exact threshold is the only way to compare the two
methods' yes/no calls on equal footing. **Everywhere else in this document — and in
the app — the ordinary reporting threshold is the more familiar
$P(\text{favourable}) > 0.95$.** This one table is the sole exception, and only
because it is pinned to the original's two-sided test.

# A worked example on real data: the microdosing trial

The natural first example is the very trial the method was built for. Following the
paper, we use only each participant's first-week, single-timepoint record — one row
per person.

(One small discrepancy, noted for honesty: the public data give an overall guess rate
of 0.647, whereas the paper's published figure annotates 0.72. The original plotting
code hard-codes 0.72 as a label; our self-test places the real unadjusted effect at
the data-derived 0.647. We flag it only because the observed-CGR identity is exactly
the kind of check that surfaces such mismatches.)

The table below runs all four outcomes through the adjustment. Read each row as: the
raw effect, the effect after imagining perfect blinding, a 95% credible interval for
that adjusted effect, and the two probabilities defined earlier.

```{r s-empirical}
emp <- do.call(rbind, lapply(names(sb), function(nm) {
  d <- sb[[nm]]; f <- cgrc(d, n_draws = N_DRAWS, seed = 2)
  h <- cgrc_headline(d, n_draws = N_DRAWS, seed = 2); s <- f$summary
  data.frame(outcome = nm, n = nrow(d), observed_CGR = round(f$observed_cgr, 3),
             `raw` = round(s$post_mean[1], 2), `adjusted (0.50)` = round(s$post_mean[2], 2),
             `95% CrI` = sprintf("[%.2f, %.2f]", s$cri_lo[2], s$cri_hi[2]),
             `P(fav)` = sprintf("%.0f%%", 100 * s$p_favourable[2]),
             `P(meaningful)` = sprintf("%.0f%%", 100 * h$p_meaningful_blind),
             check.names = FALSE)
}))
knitr::kable(emp, caption = "CGR-adjusted results, four outcomes (adjusted = perfect blinding, CGR 0.50; meaningful = beyond a half-SD difference)")
```

The figure below is the CGRC for PANAS. The horizontal axis is the assumed
correct-guess rate; the black dashed line marks perfect blinding (0.50) and the green
dashed line marks the rate actually observed. The top panel shows the estimated
effect with its credible band; the bottom panel shows the probability the effect is
favourable, at each guess rate.

```{r s-panas-fig, fig.height=6}
plot(fit_panas, title = "PANAS, week 1")
```

Now read the table and figure together, because this is where the CGRC earns its keep.
For **PANAS**, the raw effect of about
`r sprintf("%.1f", fit_panas$summary$post_mean[1])` points falls to roughly
`r sprintf("%.1f", fit_panas$summary$post_mean[2])` under perfect blinding — a drop of
about two-thirds — and, just as important, the adjusted 95% credible interval
(`r sprintf("[%.1f, %.1f]", fit_panas$summary$cri_lo[2], fit_panas$summary$cri_hi[2])`)
now straddles zero, so the probability of even a favourable effect falls to about
`r sprintf("%.0f%%", 100 * fit_panas$summary$p_favourable[2])`. In plain terms: the
PANAS result is highly sensitive to the CGRC adjustment — after reweighting to a 50%
guess rate, the posterior provides substantially weaker support for a favourable
effect.

The other three outcomes show that the adjustment is not a blunt instrument that simply
shrinks everything (all four sets of numbers are in the table above):

- **Mood VAS** behaves like PANAS — a large raw effect that roughly halves and loses its
  footing, its adjusted interval crossing zero.
- **Energy VAS** is the survivor: even after adjustment its effect stays clearly on the
  favourable side — its adjusted interval sits entirely above zero and its
  P(favourable) is essentially 100%. If any of these subjective outcomes reflects
  something beyond expectancy, this is the strongest candidate.
- **Cognitive performance** is flat with or without adjustment: near zero throughout,
  with P(favourable) barely above a coin flip. There is no effect for unblinding to
  inflate, and the method correctly leaves it unchanged — a reassuring null outcome.

None of this proves *why* the mood-type effects shrink. What it shows is that they are
**fragile to the blinding assumption** in a way the energy effect, and the (absent)
cognitive effect, are not — which is precisely the judgement the CGRC exists to
support.

# Stress-testing the method: how well does it perform for *this* design?

This is, in our view, the most important section — the one that turns the CGRC from a
formula into something you can *decide whether to rely on*. And it follows the paper's
own advice: Szigeti et al. explicitly recommended that researchers simulate how the
adjustment behaves for the specifics of their own study before applying it.

The idea behind a **simulation study** is simple and powerful: invent thousands of
fake trials in which *we already know the true answer*, run the method on each, and
tally up how well it does. The tallies are called **operating characteristics**. The
ones that matter here:

- **Bias** — on average, does the method land on the true effect, or systematically
  above/below it? (Near zero is good.)
- **Coverage** — the method's "95% credible interval" should actually contain the
  true effect about 95% of the time. Does it? (Near 0.95 is good.)
- **RMSE** — a summary of typical estimation error (smaller is better).
- **False-alarm rate** vs **detection rate** — how often it flags an effect when
  there truly is none (should be low), versus when there truly is one (should be
  high).

## The main stress test

Reusing the same 500-trials-per-scenario run from the previous section, across four
scenarios: no effect, a genuine drug effect, an expectancy confound only, and both.
The **unadjusted flag** is the probability of calling a favourable effect from the
*raw* comparison; the **adjusted flag** is the same thing from the *CGR-adjusted*
estimate — both using the standard $P > 0.95$ bar.

```{r s-sim-main}
knitr::kable(data.frame(
  scenario = scen,
  `true Delta` = round(op$true, 1),
  `adj bias` = round(op$adj_bias, 3),
  RMSE = round(op$adj_rmse, 2),
  `95% coverage` = round(op$coverage95, 3),
  `adjusted flag (P>0.95)` = round(op$p_fav_gt_95, 3),
  `unadjusted flag (P>0.95)` = round(op$unadj_p_fav_gt_95, 3),
  check.names = FALSE),
  caption = "Operating characteristics, n = 230, CGR = 0.7, 500 trials/scenario")
```

What this shows, *under this simulation model*:

- The adjusted estimate is **essentially unbiased** in every scenario, and its 95%
  credible interval covers the true adjusted effect close to 95% of the time under this
  simulation model — indicating approximately nominal interval calibration.
- When the true adjusted effect is genuinely zero (rows 1 and 3), the adjusted method
  **rarely raises a false alarm**. This is the headline win: look at the
  expectancy-only row (row 3), where the raw analysis flags an "effect" almost every
  time (0.868) purely from belief — the adjusted analysis correctly stays quiet
  (0.048).
- That protection has a **price: adjustment costs precision.** In the two rows where a
  real effect *does* exist (rows 2 and 4), the adjusted flag rate sits *below* the
  unadjusted one in the same row. This holds even in row 2, where there is *no*
  expectancy effect at all: re-estimating the contrast under a counterfactual guess
  distribution relies on the treatment × guess strata rather than the simple
  randomized-arm comparison, so the estimate is more variable and the flag rate falls
  even when activated expectancy bias is absent. When expectancy bias *is* present
  (row 4), the adjustment additionally removes the component associated with the
  altered guess distribution. Either way, the correction is not free.
- These conclusions hold *within the simulation's assumptions*. They are evidence
  about how the estimator behaves, not proof that the CGRC's real-world causal story
  is correct.

## The same tool as a "should I even use this?" check

The simulation is also a **feasibility** check, and this is where it earns its keep.
The adjustment needs at least a few people in each of the four (arm × guess) boxes. The
two "wrong-guess" boxes are filled from the minority who guess incorrectly — roughly
the $(1-\text{CGR})$ share of participants. So when the guess rate is very high *and*
the sample is small, a wrong-guess box can come up **empty** — and then the
recalculation simply cannot be done; the answer is undefined for that trial.

This is inherently a *high-guess-rate* problem. At a typical guess rate of 0.75, about
a quarter of participants guess wrong, so the boxes almost never run dry. Near 0.95,
only about 1 in 20 guesses wrong, and with few participants those boxes empty out
constantly. To show the failure clearly we deliberately pick a punishing design — 50
participants at a 95% guess rate:

```{r s-sim-sparse, cache=TRUE}
sp  <- cgr_operating(n = 50, p_cg = 0.95, mu_dte = 3, mu_aeb = 7.7, n_trials = 300, seed = 1)
r10 <- sp[sp$DTE == 1 & sp$AEB == 0, ]
knitr::kable(data.frame(design = "n = 50, CGR = 0.95",
  `empty-stratum rate` = sprintf("%.0f%%", 100 * r10$empty_stratum_rate),
  `adjusted bias (feasible trials)` = round(r10$adj_bias, 2),
  `95% coverage` = round(r10$coverage95, 3), check.names = FALSE),
  caption = "Sparse-stratum diagnostic: a thin discordant stratum leaves many trials undefined")
```

In a design like this — few people, near-certain guessing — the CGRC should **not**
be applied uncritically: about **half** the trials have an empty box and no defined answer
at all, and even the trials that *do* produce an answer have degraded interval
coverage. That is exactly why this tool is not just a validation exercise for one
paper: it is something a researcher should run **before** committing to a CGRC
adjustment on their own study.

## Running it on your own design

You supply your own sample size and expected guess rate, and read back the same
operating characteristics — bias, error, interval coverage, false-alarm behaviour, and
how often sparse boxes leave the adjustment undefined:

```{r s-tool-code, eval=FALSE}
# Supply a design resembling your trial and inspect the operating characteristics:
cgr_operating(n = 120, p_cg = 0.85, n_trials = 500)
```

# Handling "I don't know" answers

Real blinding questionnaires almost always offer a third choice: active, placebo, or
**"I don't know."** The original four-box method has no place for it, which forces an
uncomfortable choice — throw those participants away (discarding randomised data), or
recode them as if they had guessed a specific arm (inventing a belief they never
reported). Neither is good.

**`cgrc.bayes` adds a six-box extension** — this is our addition, *not* part of the
original Szigeti et al. method — that keeps "I don't know" (written `U`, for UNKNOWN)
as a genuine third answer:
$$
ACAC,\ ACPL,\ ACU,\ PLAC,\ PLPL,\ PLU.
$$
The "what if" is adjusted accordingly: we imagine the *directional* guesses (the
people who committed to active or placebo) becoming no better than chance, while
leaving the *rate of "I don't know" answers untouched* — because uncertainty is a
real, informative response, not an error to be corrected. In symbols the target is
$\Delta(0.50,\ u_{obs})$, where $u_{obs}$ is the observed "don't know" rate. Two
reassurances: if nobody answers "don't know" ($u=0$), this reduces *exactly* to the
original four-box method; and the same self-test identity holds — at the observed
directional guess rate and observed "don't know" rate, the adjusted curve lands
exactly on the raw effect.

Here it is on a simulated three-answer trial (the "identity error" near zero confirms
the self-test passes):

```{r s-unknown, cache=TRUE}
set.seed(3)
du  <- sim_aeb_unknown(n = 240, p_cg = 0.7, u = 0.25, dte_on = TRUE)   # a simulated 3-arm-guess trial
uf  <- cgrc_unknown(du, n_draws = 8000, direction = 1, seed = 1)
cnt <- lengths(cgr_unknown_strata(du))[UNKNOWN_STRATA]
zu  <- cgr_unknown_reference_line_test(du)
knitr::kable(data.frame(t(cnt), check.names = FALSE),
  caption = sprintf(
    "Six strata (simulated): observed directional CGR = %.3f, UNKNOWN rate = %.0f%%; identity error = %.1e",
    uf$observed_directional_cgr, 100 * uf$observed_unknown_rate,
    abs(zu$D_at_obs - zu$raw_mean_diff)))
```

To be clear about what this is *not*: it does **not** assume the "don't know"
participants secretly guessed wrong. Their uncertainty is kept as its own category,
neither discarded nor invented. The next section applies this to a real trial where
about a third of participants actually answered "I don't know."

# Does it hold up on other trials?

The method was not tuned to a single dataset. The three independent trials below —
spanning psychedelics, a sham dental therapy, and ketamine masked by anaesthesia — are
each run through **exactly the pipeline the app's *Analyse your own trial* panel
uses**: check the data, automatically detect whether guesses are two-way or include
"I don't know," then run the four-box or six-box analysis accordingly. They were
chosen for the lesson each teaches, not to re-report three papers.

| Trial | Design | Guess structure | Why it is useful here |
|---|---|---|---|
| Cavanna et al. (2022) | psilocybin microdosing | binary (AC/PL) | the standard four-box workflow on another psychedelic dataset |
| Santana-Penín et al. (2023) | sham dental therapy | AC / PL / UNKNOWN | ~1/3 answered "I don't know" — shows why keeping UNKNOWN matters |
| Lii et al. (2023) | ketamine masked by anaesthesia | AC / PL / UNKNOWN | directional guessing near chance — when guessing is uninformative, the correction should be tiny |

```{r s-examples, cache=TRUE}
cav <- read_ext("cavanna_cgrc.csv")
san <- read_ext("santana_cgrc_unk.csv")
lii <- read_ext("ketamine_cgrc_unk.csv")
A <- list(
  Cavanna  = app_analyse(cav, "treatment_received", "treatment_guessed", "vas_dose_days",           direction =  1),
  Santana  = app_analyse(san, "treatment_received", "treatment_guessed", "pain_improvement_6m_itt", direction =  1, delta = 1.5),
  Lii      = app_analyse(lii, "treatment_received", "treatment_guessed", "madrs_mean_day1_3",       direction = -1))
row_of <- function(nm, a) {
  s <- a$fit$summary
  ocg <- if (a$mode == "unknown") a$fit$observed_directional_cgr else a$fit$observed_cgr
  pct <- function(p) sprintf("%.0f%%", 100 * p)
  data.frame(trial = nm, mode = a$mode, n = nrow(a$trial),
             `obs CGR` = round(ocg, 3),
             raw = round(s$post_mean[1], 2), `adjusted (0.50)` = round(s$post_mean[2], 2),
             `adj 95% CrI` = sprintf("[%.2f, %.2f]", s$cri_lo[2], s$cri_hi[2]),
             `P(fav) raw->adj` = paste(pct(s$p_favourable[1]), "->", pct(s$p_favourable[2])),
             `P(meaningful) raw->adj` = paste(pct(a$head$p_meaningful_obs), "->", pct(a$head$p_meaningful_blind)),
             check.names = FALSE)
}
knitr::kable(do.call(rbind, Map(row_of, names(A), A)), row.names = FALSE,
  caption = "Three independent trials through the app's analysis pipeline (Cavanna outcome = dosing-day VAS, a high-signal outcome)")
```

How to read each row: compare the **raw** and **adjusted (0.50)** columns to see how far
the "what if blinding had held?" recalculation moves the estimate, then use the two
probability columns — *is there any effect* (P(fav)) and *is it big enough to matter*
(P(meaningful)) — to judge what survives. The three trials land in three different
regimes.

**Cavanna provides a useful high-signal example.** The outcome is a dosing-day sensation
rating — something people can plausibly feel — so a large effect is expected. Guessing
sits only a little above chance
(`r sprintf("%.3f", A$Cavanna$fit$observed_cgr)`), so the adjustment barely touches the
estimate (`r sprintf("%.1f", A$Cavanna$fit$summary$post_mean[1])` →
`r sprintf("%.1f", A$Cavanna$fit$summary$post_mean[2])`). The effect is large but, with
only `r nrow(A$Cavanna$trial)` participants, uncertain — its adjusted credible interval,
`r sprintf("[%.1f, %.1f]", A$Cavanna$fit$summary$cri_lo[2], A$Cavanna$fit$summary$cri_hi[2])`,
is wide and dips below zero. The lesson is reassuring: a large effect is not erased by
the correction; it is simply reported with a correspondingly wide credible interval.

**Santana is the most instructive case, and it cuts two ways.** On the one hand the
estimated pain-improvement effect remains strongly positive after adjustment —
P(favourable) stays
at `r sprintf("%.0f%%", 100 * A$Santana$fit$summary$p_favourable[2])`, and the adjusted
interval,
`r sprintf("[%.2f, %.2f]", A$Santana$fit$summary$cri_lo[2], A$Santana$fit$summary$cri_hi[2])`,
sits entirely above zero. Whether it is *clinically meaningful* is the genuine open
question: measured against the trial's 1.5-point threshold, the probability slips from
about `r sprintf("%.0f%%", 100 * A$Santana$head$p_meaningful_obs)` to
`r sprintf("%.0f%%", 100 * A$Santana$head$p_meaningful_blind)` under adjustment — from a
hair above even odds to a hair below — because the adjusted estimate
(`r sprintf("%.2f", A$Santana$fit$summary$post_mean[2])`) lands just under the bar. On
the other hand, Santana is *also* a live example of the feasibility caution from the
stress-test section. Keeping the
`r sprintf("%.0f%%", 100 * A$Santana$fit$observed_unknown_rate)` of participants who
answered "I don't know" is exactly why all `r nrow(A$Santana$trial)` of them stay in the
analysis — but so many of the rest guessed the same way that some of the discordant boxes
hold only a handful of people (as few as
`r as.integer(lengths(cgr_unknown_strata(A$Santana$trial))["ACPL"])`). The adjusted
number here is therefore best read as *defined but fragile* — precisely the situation the
feasibility panel is built to flag.

**Lii is the near-chance directional-guessing case**, and shows the correction going
quiet when directional guessing carries little information. Directional guessing is
essentially at chance
(`r sprintf("%.3f", A$Lii$fit$observed_directional_cgr)`, against 0.50 for a coin flip),
so the adjustment is almost inert: the antidepressant signal moves only from
`r sprintf("%.2f", A$Lii$fit$summary$post_mean[1])` to
`r sprintf("%.2f", A$Lii$fit$summary$post_mean[2])` on the MADRS depression scale (where
a *negative* change means improvement), and P(favourable) barely changes
(`r sprintf("%.0f%%", 100 * A$Lii$fit$summary$p_favourable[1])` →
`r sprintf("%.0f%%", 100 * A$Lii$fit$summary$p_favourable[2])`). The effect itself is
uncertain — its interval crosses zero with only `r nrow(A$Lii$trial)` participants — but
that is a *sample-size* limitation, not an unblinding one, and its six boxes are
comfortably filled precisely because at-chance guessing spreads people evenly across them.

Taken together, the adjustment behaves as intended across three different regimes: it
leaves a large effect intact (Cavanna), tempers a real-but-borderline effect while
flagging its own fragility (Santana), and stays close to the unadjusted estimate when
directional guessing is near chance (Lii). The two curves below make the Santana and Lii
cases visual:

```{r s-examples-plot, fig.height=5.5, fig.show="hold", out.width="49%"}
app_curve(A$Santana, "favourable")   # UNKNOWN preserved; effect at the clinical MID
app_curve(A$Lii,     "favourable")   # near-chance guessing; correction is small
```

The same workflow behaves sensibly across very different trials. Full provenance and
data-preparation notes for each dataset are documented with the data and in the
companion technical document.

# The app: putting all of it in a researcher's hands

Everything above was run through the *same* analysis logic the app exposes — the
worked examples are not one-off scripts. The point of the app is to let a researcher
who did **not** develop this method, and does **not** want to write R code, get the
full analysis and its design check by pointing and clicking.

## Panel 1 — Analyse your own trial

You upload a spreadsheet with one row per participant and three columns: which arm they
were in, which arm they guessed, and their outcome. The app then:

1. **checks** the data (missing values, unrecognised codes, how many people fall in
   each box, the guess rate in each arm);
2. **detects automatically** whether the guesses are two-way or include "I don't know";
3. runs the four-box or six-box analysis accordingly;
4. draws the **CGRC curve** (the same kind of figure shown earlier);
5. reports the **raw** and **adjusted** effect with a 95% credible interval;
6. reports the two plain-language probabilities — *is there an effect* and *is it big
   enough to matter*.

## Panel 2 — Design check: operating characteristics for your design

The second panel runs the simulation stress-test for a design you describe, and hands
back the operating-characteristics read-out — bias, interval coverage, false-alarm
behaviour, and how often sparse boxes make the answer undefined — as a plain-language
summary. Together, the two panels are the whole point:

> The first panel tells you **what the CGRC adjustment says** for your trial; the
> second tells you **whether the adjustment has acceptable operating characteristics**
> for a design like yours.

## Try it

Under the hood the whole app is a thin wrapper over a few one-line function calls, so
you can do the same thing from code if you prefer:

```{r s-app-code, eval=FALSE}
# 1. Adjust a trial (df has columns condition = AC/PL, guess = AC/PL[/UNKNOWN], value):
cgrc(my_trial)                                  # or cgrc_unknown(my_trial) when UNKNOWN is present

# 2. Check the adjustment's operating characteristics for a design like yours:
cgr_operating(n = 120, p_cg = 0.85, n_trials = 500)

# 3. Do both interactively, no code required:
install.packages("shiny")
cgrc.bayes::cgrc_app()                          # launches the "Design" + "Analyse your own trial" app
```

`cgrc_app()` opens the interactive tool on your own machine (see the repository README
for launch details); there is no hosted website version. It is the most direct way to
run the whole workflow — analysis and design check together — on your own
data.

# What this does and does not claim

It helps to separate the pieces cleanly.

- **What stays the same as the original.** We compute exactly the same CGRC target,
  $\Delta(c)$, and satisfy the same self-test identity. We did not redefine what is
  being estimated.
- **What our Bayesian version adds.** A directly interpretable posterior uncertainty
  summary — a credible interval and the two plain-language probabilities — in place of
  an averaged p-value. (Our version still uses random draws, so it is not perfectly
  exact either; but those draws *summarise a well-defined answer* rather than
  *construct* the estimate, and adding more draws simply sharpens the picture toward a
  fixed target.)
- **What the simulation adds.** A way to see, before you apply it, whether the
  adjustment is roughly unbiased, approximately calibrated, and even computable for
  your particular design — exactly the "simulate first" step the original paper
  recommended.
- **What the "I don't know" extension adds.** A principled way to keep genuine "don't
  know" answers instead of discarding or inventing them — a fix to how the data are
  handled, clearly labelled as our addition.
- **What the worked examples add.** Evidence that one workflow behaves sensibly across
  trials where blinding was poor, mixed, or largely successful.
- **What the app adds.** Access. A trial researcher should not have to re-derive the
  mathematics or write custom code to ask whether unblinding could plausibly be
  distorting their result — and whether the correction is even stable for their design.
  (Keeping the full posterior also enables "is the effect practically negligible?"
  analyses, demonstrated in the full technical document.)

**And, most importantly, what none of this establishes.** The defensible claim is
narrow: *we rebuilt the same CGRC estimate with directly interpretable uncertainty,
characterised its operating behaviour by simulation, extended it to real "don't know"
answers, and packaged it so others can use it.* In particular:

- The shrinkage under adjustment is **not** a clean measure of "how much of the effect
  was caused by expectancy."
- Guessing happens *after* randomisation and can itself reflect real drug effects or
  side effects, which the re-weighting cannot untangle.
- Reproducing the original validates the *computation*, not its underlying causal
  assumptions.
- The simulations validate behaviour *within the simulation's model*, not real-world
  causal truth.
- The "don't know" extension fixes a data-handling problem, not the deeper causal
  limits of the method.

The bottom line: the CGRC — in either the original or our version — is best understood
as a **sensitivity analysis for unblinding and expectancy**, a way to ask how fragile
a result is to broken blinding, not a causal decomposition of the treatment effect.

# Sources

Only the first two references underpin this work; full per-finding citations are in
`reports/SOURCES.md`, and the external-trial datasets are cited with their archives in
`inst/extdata/external_studies/PROVENANCE.txt`.

- **Primary methodological source** — the original CGRC method reproduced here.
  Szigeti B, Nutt D, Carhart-Harris R, Erritzoe D. 2023. "The difference between
  'placebo group' and 'placebo control': a case study in psychedelic microdosing."
  *Scientific Reports* 13:12107. <https://doi.org/10.1038/s41598-023-34938-7>
- **Empirical dataset** — source of the public week-1 data reanalysed here.
  Szigeti B, et al. 2021. "Self-blinding citizen science to explore psychedelic
  microdosing." *eLife* 10:e62878. <https://doi.org/10.7554/eLife.62878>

Worked-example datasets: Cavanna et al. (2022) *Transl. Psychiatry* 12:307; Santana-Penín
et al. (2023) *Ann. Anat.* 250:152117; Lii et al. (2023) *Nat. Mental Health* 1:876–886.
This project reproduces and re-analyses published work; it is independent of and not
endorsed by the original authors.

```{r s-session, echo=FALSE}
sessionInfo()
```
